\documentclass[11pt]{article}

% ── Packages ──────────────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{float}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{multirow}
\usepackage{array}
\usepackage{listings}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\pgfplotsset{compat=1.18}
\usepackage{natbib}
\usepackage{authblk}
% algorithm package not available on this system; using listings for pseudocode
\usepackage{tcolorbox}
\usepackage{longtable}
\usepackage{rotating}
\usepackage{afterpage}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning, fit, backgrounds}

% ── Colors ────────────────────────────────────────────────────────────────────
\definecolor{sparkblue}{HTML}{58a6ff}
\definecolor{sparkgreen}{HTML}{3fb950}
\definecolor{sparkorange}{HTML}{d29922}
\definecolor{sparkred}{HTML}{f85149}
\definecolor{sparkpurple}{HTML}{bc8cff}
\definecolor{sparkgray}{HTML}{8b949e}
\definecolor{lightgray}{HTML}{f6f8fa}

\hypersetup{
  colorlinks=true,
  linkcolor=sparkblue,
  citecolor=sparkgreen,
  urlcolor=sparkblue,
}

% ── Listing style ─────────────────────────────────────────────────────────────
\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  rulecolor=\color{gray!30},
  backgroundcolor=\color{gray!5},
  keywordstyle=\color{sparkblue}\bfseries,
  commentstyle=\color{sparkgray}\itshape,
  stringstyle=\color{sparkgreen},
  numbers=left,
  numberstyle=\tiny\color{sparkgray},
  numbersep=5pt,
}

% ── Theorem environments ──────────────────────────────────────────────────────
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}

% ── TikZ styles ───────────────────────────────────────────────────────────────
\tikzset{
  stage/.style={rectangle, rounded corners=3pt, draw=sparkblue!70, fill=sparkblue!10,
                text width=6.5cm, minimum height=0.55cm, align=center, font=\small},
  stagered/.style={stage, draw=sparkred!70, fill=sparkred!8},
  stagegreen/.style={stage, draw=sparkgreen!70, fill=sparkgreen!8},
  stageorange/.style={stage, draw=sparkorange!70, fill=sparkorange!8},
  stagepurple/.style={stage, draw=sparkpurple!70, fill=sparkpurple!8},
  arrow/.style={-{Stealth[length=4pt]}, thick, color=sparkgray},
}

% ── Title ──────────────────────────────────────────────────────────────────────
\title{%
  \textbf{memory-spark: GPU-Accelerated Persistent Memory for Autonomous AI Agents}\\[0.5cm]
  \large A Production System for Agentic Retrieval-Augmented Generation\\
  with Dynamic Reranker Gating, Reciprocal Rank Fusion,\\
  and Multi-Pool Semantic Memory
}

\author[1]{Klein Panic}
\affil[1]{Independent Researcher, \texttt{klein@exampleuser.com}}

\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
We present \textbf{memory-spark}, a production memory substrate for autonomous AI agents
that continuously ingests workspace knowledge, indexes it with hybrid dense--sparse
retrieval, and injects high-relevance context before each conversational turn.
Unlike standard Retrieval-Augmented Generation (RAG) systems designed for single-turn
question answering, memory-spark is engineered for the fundamentally harder problem of
persistent agentic memory: heterogeneous source types (workspace files, tool schemas,
mistake logs, reference documents, captured preferences), strict sub-second latency
requirements (memory injected before \emph{every} turn, not just user queries),
cross-agent knowledge sharing with data isolation, and active defense against
self-poisoning from agent-generated errors.

The system operates entirely on local GPU infrastructure (NVIDIA DGX Spark) with no
cloud API dependencies for embeddings, reranking, or LLM inference. We introduce three
key technical contributions: (1)~a \textbf{Dynamic Reranker Gate} that analyzes
first-stage vector score distributions to intelligently bypass cross-encoder reranking
when retrieval confidence is already high, reducing latency by $\sim$50\% with no
quality loss; (2)~\textbf{scale-invariant Reciprocal Rank Fusion} (RRF) for hybrid
dense--sparse merging and retrieve--rerank blending, resolving the fundamental
incompatibility between BM25 score ranges (5--20+) and cosine similarity ranges
(0.2--0.6) that caused catastrophic rank-washout in prior score-based approaches; and
(3)~a \textbf{15-stage retrieval pipeline} with instruction-aware asymmetric
embeddings, multi-pool search, source priority weighting, temporal decay,
Hypothetical Document Embeddings (HyDE), multi-query expansion, Maximal Marginal
Relevance (MMR) diversity, parent-child chunk expansion, and layered security filtering.

We evaluate on the BEIR benchmark \citep{thakur2021beir} across three datasets.
On SciFact (300 queries), our production configuration GATE-A achieves NDCG@10 of
0.7802, Recall@10 of 0.9137, and mean latency of 626ms while invoking the cross-encoder
on only 21\% of queries. Our best configuration (Config~U, logit blend $\alpha=0.4$)
achieves NDCG@10 of 0.7889 and Recall@10 of 0.9099. On FiQA (648 queries, financial
Q\&A), GATE-A achieves 0.5479 NDCG@10 (+66\% over BM25, +68\% over Contriever). On
NFCorpus (323 queries, cross-domain medical retrieval), Vector-Only achieves 0.4443
NDCG@10 (+35.5\% over Contriever), while reranking \emph{hurts} performance due to the
cross-domain query format mismatch.

A detailed forensic analysis of the 12-phase development cycle documents eight
categories of production failure modes, including silent NaN propagation via Apache Arrow
type mismatches, BM25 sigmoid score saturation causing rank-washout, cross-encoder score
compression, and pipeline stage ordering bugs. We argue that the gap between benchmark
RAG and production agentic RAG is larger than commonly acknowledged, and that iterative
quantitative evaluation (not intuition) is the only reliable guide to system improvement.

The complete system serves the meta agent in production with 18 self-service
memory tools, 37,000+ indexed chunks, and a 483-test validation suite. All code,
benchmarks, and evaluation scripts are open-source at
\url{https://github.com/exampleuser/memory-spark}.
\end{abstract}

% ══════════════════════════════════════════════════════════════════════════════
\section{Introduction}
\label{sec:intro}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{The Persistent Memory Problem for Autonomous Agents}

Long-running AI agents face a memory problem that is categorically different from the
context management challenges of single-turn assistants. A conversational assistant
must remember what was said ten messages ago. An autonomous agent running 24 hours a
day, executing dozens of tasks per session across hundreds of sessions, must remember
what was learned \emph{weeks} ago---that a particular configuration key must never be
modified during live operation, that a previous attempt at a seemingly safe change
caused a production outage, that a specific file path is sacred, that a colleague
prefers concise responses over detailed reports.

The context window of even the largest LLMs (128k--1M tokens) is insufficient for this
task, and for good reason: naïvely injecting everything is both token-expensive and
architecturally fragile. A 128k token context filled with agent logs leaves little
room for the actual task. More fundamentally, irrelevant context degrades performance
\citep{shi2023large}: a model trying to use 50,000 tokens of historical workspace
content to answer a simple question about a configuration value performs worse than one
given the 200 most relevant tokens.

Retrieval-Augmented Generation (RAG) \citep{lewis2020rag} addresses this by retrieving
relevant documents at query time. But production agentic RAG introduces challenges
absent from standard QA benchmarks:

\begin{description}[leftmargin=1.5em, labelindent=0em, itemsep=4pt]
  \item[\textbf{Source heterogeneity.}] Agent workspaces contain markdown documentation,
  TypeScript source code, tool schemas, daily memory notes, mistake logs with root cause
  analyses, captured user preferences, reference PDFs, and session transcripts. Each
  source type has different retrieval characteristics, different relevance signals, and
  different weighting requirements. A mistake entry from three months ago that is
  \emph{exactly} applicable to the current task should outrank a generic document that
  scores slightly higher on cosine similarity.

  \item[\textbf{Latency constraints.}] In production, memory retrieval happens before
  \emph{every} agent turn---not just when the user explicitly asks a factual question.
  If retrieval takes 3 seconds, every conversational exchange has a 3-second tax. This
  is unacceptable for interactive use. Sub-second latency is a hard requirement.

  \item[\textbf{Temporal dynamics.}] Recent information should generally outweigh old
  information, but not uniformly. A user preference captured yesterday is more reliable
  than one from six months ago. But a safety rule (``never edit openclaw.json during a
  heartbeat run'') is timeless and should not be temporally discounted.

  \item[\textbf{Cross-agent sharing.}] A multi-agent system where agents can share
  knowledge but must maintain data isolation presents a design tension. Agent~A's
  per-session scratchpad should not leak into Agent~B's context, but a mistake entry
  marked ``shared'' should be visible to all agents.

  \item[\textbf{Self-poisoning.}] Agents can capture their own mistakes. An agent that
  confidently (but incorrectly) states a fact, captures that statement as a ``fact'',
  and later retrieves it to reinforce the incorrect belief creates a feedback loop that
  is difficult to detect and correct. The memory system must actively prevent this.

  \item[\textbf{Cold infrastructure.}] Production systems run on hardware with variable
  availability. If the embedding service is temporarily unavailable, the entire agent
  pipeline should degrade gracefully rather than permanently initializing into a broken
  state.
\end{description}

\subsection{Research Questions}

This paper addresses four research questions arising from these challenges:

\begin{enumerate}[leftmargin=1.5em, itemsep=4pt]
  \item[\textbf{RQ1.}] Can a cross-encoder reranker be used \emph{selectively}---only
  when it will actually help---to simultaneously improve recall and reduce latency
  compared to unconditional reranking?

  \item[\textbf{RQ2.}] How can dense and sparse retrieval results be fused when their
  score scales are fundamentally incompatible, and how does this choice affect the
  complete retrieve--rerank pipeline?

  \item[\textbf{RQ3.}] What failure modes are endemic to production agentic RAG that do
  not appear in standard benchmark evaluations, and how can they be systematically
  detected and remediated?

  \item[\textbf{RQ4.}] How does retrieval pipeline architecture affect quality across
  qualitatively different BEIR datasets, and what does this reveal about the conditions
  under which reranking helps vs.\ hurts?
\end{enumerate}

\subsection{Contributions}

We make the following contributions:

\begin{itemize}[leftmargin=1.5em, itemsep=3pt]
  \item \textbf{Dynamic Reranker Gate (Section~\ref{sec:gate})}: A threshold-based gate
  that uses the spread of top-$k$ vector similarity scores to decide whether
  cross-encoder reranking will provide meaningful lift. On BEIR SciFact, this reduces
  reranker invocations by 79\% while improving Recall@10 by 1.1\% vs.\ unconditional
  reranking (answering RQ1).

  \item \textbf{Scale-Invariant Hybrid Fusion via RRF (Section~\ref{sec:rrf})}: We apply
  Reciprocal Rank Fusion \citep{cormack2009reciprocal} to both the dense--sparse hybrid
  merge and the retrieve--rerank blending steps, eliminating the need for score
  normalization and resolving catastrophic rank-washout caused by BM25/cosine scale
  incompatibility (answering RQ2).

  \item \textbf{Production Failure Mode Taxonomy (Section~\ref{sec:failures})}: An
  empirical catalog of eight failure categories discovered during a 12-phase iterative
  development cycle, including silent failures from type system mismatches, score
  distribution pathologies, security vulnerabilities, and pipeline ordering errors
  (answering RQ3).

  \item \textbf{Cross-Dataset Retrieval Analysis (Section~\ref{sec:eval})}: A systematic
  structural analysis of why reranking helps on SciFact and FiQA but hurts on NFCorpus,
  grounded in query/corpus format differences and reranker training distribution
  (answering RQ4).

  \item \textbf{Production System}: A complete, deployed memory substrate serving 6
  autonomous agents with 18 plugin tools, documented engineering lessons, and
  open-source implementation.
\end{itemize}

\subsection{Paper Organization}

Section~\ref{sec:background} establishes mathematical preliminaries for all retrieval
components. Section~\ref{sec:arch} describes the complete system architecture.
Sections~\ref{sec:gate} and~\ref{sec:rrf} present the two primary technical
innovations in depth. Section~\ref{sec:hyde} analyzes HyDE. Section~\ref{sec:failures}
catalogs production failure modes. Section~\ref{sec:eval} presents experimental results.
Section~\ref{sec:deployment} covers production engineering. Section~\ref{sec:related}
discusses related work. Section~\ref{sec:future} outlines future directions.
Section~\ref{sec:conclusion} concludes.

% ══════════════════════════════════════════════════════════════════════════════
\section{Background and Preliminaries}
\label{sec:background}
% ══════════════════════════════════════════════════════════════════════════════

We establish formal definitions for the retrieval components used throughout this paper.

\subsection{Dense Retrieval: Bi-Encoder Formulation}

A bi-encoder \citep{karpukhin2020dpr} maps queries and documents into a shared embedding
space using two (possibly identical) encoders:

\begin{align}
  \mathbf{q} &= f_Q(q) \in \mathbb{R}^d \\
  \mathbf{d} &= f_D(d) \in \mathbb{R}^d
\end{align}

where $f_Q$ and $f_D$ are the query and document encoders respectively, and $d$ is the
embedding dimension ($d=4096$ in our system). Relevance is scored via cosine similarity:

\begin{equation}
  \text{sim}(q, d) = \frac{\mathbf{q} \cdot \mathbf{d}}{\|\mathbf{q}\|\,\|\mathbf{d}\|}
  \in [-1, 1]
\end{equation}

In practice, after $\ell_2$ normalization, this reduces to the dot product
$\mathbf{q} \cdot \mathbf{d}$.

\paragraph{Asymmetric (Instruction-Aware) Encoding.}
Instruction-tuned embedding models \citep{su2022instructor} produce better
query--document alignment when queries include task-specific instruction prefixes while
documents are embedded without them. We enforce:

\begin{align}
  \mathbf{q} &= f\!\left(\text{``Instruct: } t \text{\textbackslash nQuery: } q\text{''}\right) \\
  \mathbf{d} &= f(d)
\end{align}

where $t$ is the retrieval task description (``Given a question, retrieve relevant
passages that answer the query''). Omitting the prefix for queries causes $\mathbf{q}$
to land in the document subspace, degrading cosine similarity discrimination by
moving queries into the document region of the embedding space where they were never
trained to reside.

\paragraph{IVF\_PQ Approximate Nearest Neighbor Search.}
Exact nearest neighbor search over $N$ vectors costs $O(N \cdot d)$. For large corpora,
we use IVF\_PQ (Inverted File with Product Quantization) \citep{johnson2019faiss}:

\begin{enumerate}[leftmargin=2em, itemsep=2pt]
  \item \textbf{IVF}: Partition the vector space into $P$ clusters (Voronoi cells) via
  $k$-means. At query time, search only the $n_{\text{probe}}$ nearest centroids.
  Search complexity reduces to $O(n_{\text{probe}} \cdot N/P \cdot d)$.
  \item \textbf{PQ}: Decompose each vector into $M$ sub-vectors of dimension $d/M$,
  independently quantizing each sub-vector to $K=256$ centroids. Storage reduces from
  $d \cdot 32$-bit floats to $M$ bytes per vector.
\end{enumerate}

Our configuration: $P=10$ IVF partitions, $M=64$ PQ sub-vectors, cosine metric.
This trades a small quality penalty ($\approx$1--2\% NDCG on our corpus) for $\sim$50$\times$
faster search.

\subsection{Sparse Retrieval: BM25 Formulation}

BM25 \citep{robertson1976relevance} scores a document $D$ for query $Q$ as:

\begin{equation}
  \text{BM25}(D, Q) = \sum_{t \in Q} \text{IDF}(t) \cdot
  \frac{f(t, D) \cdot (k_1 + 1)}{f(t, D) + k_1 \cdot \left(1 - b + b \cdot \frac{|D|}{\text{avgdl}}\right)}
\end{equation}

where $f(t, D)$ is the term frequency of term $t$ in document $D$, $|D|$ is document
length, $\text{avgdl}$ is average document length, and the IDF term:

\begin{equation}
  \text{IDF}(t) = \log\!\left(\frac{N - n(t) + 0.5}{n(t) + 0.5} + 1\right)
\end{equation}

with $N$ the corpus size and $n(t)$ the document frequency of term $t$. Typical
parameters: $k_1 = 1.2$, $b = 0.75$.

BM25 scores are unbounded above and depend on corpus statistics. In our system, BM25
scores range from 5 to 20+ for relevant documents. This scale incompatibility with
cosine similarity ($\in [0.2, 0.6]$ for relevant documents) is the root cause of
rank-washout when score-based fusion is applied without careful normalization
(Section~\ref{sec:rrf:motivation}).

We implement BM25 via LanceDB's built-in FTS module, which uses tantivy for inverted
index construction and BM25 scoring.

\subsection{Evaluation Metrics}

Let $\mathcal{R}_q$ be the set of relevant documents for query $q$, and let
$\hat{r}_1, \hat{r}_2, \ldots, \hat{r}_k$ be the ranked retrieval result (top-$k$).

\begin{definition}[NDCG@k]
Normalized Discounted Cumulative Gain at rank $k$:
\begin{equation}
  \text{NDCG@}k = \frac{\text{DCG@}k}{\text{IDCG@}k}, \quad
  \text{DCG@}k = \sum_{i=1}^{k} \frac{\text{rel}(\hat{r}_i)}{\log_2(i+1)}
\end{equation}
where $\text{rel}(\hat{r}_i)$ is the relevance grade of the item at rank $i$ (binary 0/1
in SciFact and FiQA; graded 0/1/2 in NFCorpus), and IDCG@$k$ is the DCG@$k$ of the
ideal ranking (all relevant documents ranked first). NDCG@10 is our primary metric.
\end{definition}

\begin{definition}[Recall@k]
\begin{equation}
  \text{Recall@}k = \frac{|\{\hat{r}_1,\ldots,\hat{r}_k\} \cap \mathcal{R}_q|}{|\mathcal{R}_q|}
\end{equation}
The fraction of relevant documents retrieved in the top-$k$ results. Bounded by 1.0.
Note: when $|\mathcal{R}_q| > k$, Recall@$k$ is strictly less than 1.0 under any
retrieval scheme, creating a structural ceiling that significantly affects NFCorpus
(mean $|\mathcal{R}_q| = 38.2$; maximum achievable Recall@10 $\approx 0.61$).
\end{definition}

\begin{definition}[MAP@k]
Mean Average Precision at rank $k$:
\begin{equation}
  \text{MAP@}k = \frac{1}{|Q|} \sum_{q \in Q} \text{AP@}k(q), \quad
  \text{AP@}k(q) = \frac{1}{|\mathcal{R}_q|} \sum_{i=1}^{k} \text{rel}(\hat{r}_i) \cdot \frac{|\text{relevant@}i|}{i}
\end{equation}
The denominator $|\mathcal{R}_q|$ (total relevant documents in the corpus) is standard
BEIR semantics. We note that using $\min(|\mathcal{R}_q|, k)$ as denominator instead
inflates AP@$k$ on queries with few relevant documents; this was a bug in our initial
implementation (failure mode F7, Section~\ref{sec:failures}).
\end{definition}

\begin{definition}[MRR]
Mean Reciprocal Rank:
\begin{equation}
  \text{MRR} = \frac{1}{|Q|} \sum_{q \in Q} \frac{1}{\text{rank}_q}
\end{equation}
where $\text{rank}_q$ is the rank of the first relevant document for query $q$. MRR
rewards systems that place at least one relevant result near the top, without requiring
comprehensive coverage.
\end{definition}

\subsection{Cross-Encoder Reranking}

A cross-encoder \citep{nogueira2019passage} jointly encodes the query--document pair and
outputs a relevance score:

\begin{equation}
  s_{\text{cross}}(q, d) = g(q \oplus d) \in \mathbb{R}
\end{equation}

where $\oplus$ denotes concatenation and $g$ is typically a BERT-style model with a
linear head. Cross-encoders are significantly more accurate than bi-encoders because
they can model query--document interactions at every token, but their inference cost is
$O(N)$ per query (no precomputed document representations), making them practical only
as a second-stage reranker on a small candidate set.

Our cross-encoder is \texttt{llama-nemotron-rerank-1b-v2}, a 1B-parameter model
optimized for reranking. In Section~\ref{sec:failures:F3} we document a score
compression pathology specific to this model size.

\subsection{Reciprocal Rank Fusion}

\begin{definition}[RRF \citep{cormack2009reciprocal}]
Given a set $\mathcal{L}$ of ranked lists and a smoothing constant $k > 0$, the RRF
score for document $d$ is:
\begin{equation}
  \text{RRF}(d; \mathcal{L}, k) = \sum_{L \in \mathcal{L}} \frac{w_L}{k + \text{rank}_L(d)}
\end{equation}
where $\text{rank}_L(d)$ is the position of $d$ in list $L$ (1-indexed; documents not
in $L$ are assigned rank $|\mathcal{L}_{\max}| + 1$), and $w_L$ is a per-list weight
(default $w_L = 1$ for all lists). The smoothing constant $k=60$ is the standard value
recommended by Cormack et al.
\end{definition}

RRF is scale-invariant: the absolute values of scores in each list are irrelevant; only
ranks matter. Two documents ranked 1st and 5th by vector search receive scores $1/61$
and $1/65$, regardless of whether their cosine similarities are 0.90 and 0.85, or 0.55
and 0.52. This makes RRF naturally robust to the BM25/cosine scale mismatch.

\subsection{Maximal Marginal Relevance}

\begin{definition}[MMR \citep{carbonell1998mmr}]
Let $S$ be the set of already-selected documents and $R$ the remaining candidate set.
The next document is:
\begin{equation}
  d^* = \arg\max_{d \in R} \left[\lambda \cdot \text{sim}(q, d)
    - (1-\lambda) \cdot \max_{d' \in S} \text{sim}(d, d')\right]
\end{equation}
where $\lambda \in [0,1]$ controls the diversity--relevance trade-off. High $\lambda$
($\to 1$) reproduces standard relevance ranking; low $\lambda$ ($\to 0$) maximizes
diversity. Our default $\lambda = 0.9$ applies light diversity regularization.
\end{definition}

MMR is applied as the \emph{final} pipeline stage, after reranking, so it selects
diverse documents from among those the cross-encoder has already confirmed as relevant.
Applying MMR before reranking (a prior configuration) was counterproductive: it
discarded high-relevance candidates before the reranker could confirm their quality.

We implement MMR with cosine similarity in the 4096-dimensional embedding space.
A critical implementation note (failure mode F1): Arrow Vector objects returned by
LanceDB require explicit conversion to JavaScript number arrays via \texttt{.toArray()}
before cosine similarity can be computed; bracket indexing returns \texttt{undefined}.

\subsection{Hypothetical Document Embeddings}

\begin{definition}[HyDE \citep{gao2023hyde}]
Instead of embedding the query directly, generate a hypothetical document
$\hat{d} \sim p(\cdot \mid q)$ using a language model, then embed it as a document:
\begin{equation}
  \mathbf{q}_{\text{HyDE}} = f_D(\hat{d}), \quad \hat{d} \sim \text{LLM}(q)
\end{equation}
The hypothesis: the LLM's hypothetical document lives in the same embedding subspace
as actual relevant documents (the document subspace), whereas the query $q$ may reside
in a different region (the question subspace). Replacing the query embedding with a
document-space vector bridges this gap.
\end{definition}

HyDE is theoretically motivated for tasks with large query--document distributional
gaps. In Section~\ref{sec:hyde} we analyze why it failed on BEIR in our setup and
discuss when it is likely to help.

% ══════════════════════════════════════════════════════════════════════════════
\section{System Architecture}
\label{sec:arch}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{Overview and Design Principles}

memory-spark implements a plugin for the OpenClaw autonomous agent framework. It hooks
into the agent lifecycle at three insertion points:

\begin{itemize}[leftmargin=1.5em, itemsep=3pt]
  \item \textbf{\texttt{before\_prompt\_build}}: Auto-recall. Retrieves the top-$k$ most
  relevant memory chunks and injects them into the agent's system prompt, token-budgeted
  to 2000 tokens.
  \item \textbf{\texttt{agent\_end}}: Auto-capture. Extracts facts, decisions,
  preferences, and code snippets from the completed agent turn, classifies them, and
  stores them in the appropriate memory pool.
  \item \textbf{\texttt{after\_compaction}}: Re-indexes newly compacted session
  transcripts to make session history searchable.
\end{itemize}

Seven design principles govern the system:

\begin{description}[leftmargin=1.5em, labelindent=0em, itemsep=3pt]
  \item[\textbf{No workarounds.}] Every implementation is the proper solution. Temporary
  hacks create technical debt that compounds as the system grows.
  \item[\textbf{Scientifically reproducible.}] All benchmarks produce identical results
  given identical inputs. No random seeds without documentation.
  \item[\textbf{Configurable over hardcoded.}] Every threshold, timeout, and model is
  overridable via the OpenClaw plugin config schema.
  \item[\textbf{Data isolation with shared access.}] Each agent owns its pools. Cross-agent
  access is explicit, auditable, and filter-enforced at the storage layer.
  \item[\textbf{Reference never auto-injects.}] Reference documents (PDFs, library docs)
  are retrieved only via explicit tool calls, never silently injected into context.
  \item[\textbf{Fail open, degrade gracefully.}] If Spark is unavailable at startup,
  the plugin still initializes. If embedding fails, capture is skipped but the turn
  completes.
  \item[\textbf{Quantitative validation.}] Every architectural decision has a
  corresponding BEIR benchmark measurement.
\end{description}

\subsection{Infrastructure}

All ML inference runs on a local NVIDIA DGX Spark node with no cloud API dependencies:

\begin{table}[H]
\centering
\small
\begin{tabular}{llll}
\toprule
\textbf{Component} & \textbf{Model} & \textbf{Port} & \textbf{Notes} \\
\midrule
Embeddings & nvidia/llama-embed-nemotron-8b & 18091 & 4096-dim, instruction-tuned \\
Reranker & nvidia/llama-nemotron-rerank-1b-v2 & 18096 & Cross-encoder, 1B params \\
LLM (HyDE) & NVIDIA-Nemotron-Super-120B-A12B & 18080 & OpenAI-compat API \\
OCR (VL) & zai-org/GLM-OCR & 18080 & 0.9B VL, \#1 OmniDocBench V1.5 \\
OCR (fallback) & EasyOCR & 18097 & Legacy Python service \\
STT & nvidia/Parakeet-CTC-1.1B & 18094 & Speech-to-text \\
NER & (FastAPI service) & 18112 & Named entity extraction \\
Zero-shot & (FastAPI service) & 18113 & Capture classification \\
Storage & LanceDB (IVF\_PQ + FTS) & local & Single-table architecture \\
\bottomrule
\end{tabular}
\caption{Inference stack. All services are self-hosted on a single NVIDIA DGX Spark node.
Note: The LLM (HyDE) and GLM-OCR share port 18080 via vLLM, a known scheduling conflict
addressed in the Spark v2 architecture plan (Section~\ref{sec:future}).}
\label{tab:infra}
\end{table}

\subsection{Storage Model: Single-Table LanceDB Architecture}

We use a single LanceDB table \texttt{memory\_chunks} with a \texttt{pool} column for
logical isolation. LanceDB officially recommends single-table over multi-table for
datasets under 10M rows, as it provides better partition distribution, zero-copy schema
evolution, and simpler infrastructure.

The table schema:

\begin{lstlisting}[language=bash, caption={memory\_chunks table schema}]
id:             VARCHAR  -- UUID, 16-char
text:           VARCHAR  -- chunk text, max ~2000 tokens
vector:         VECTOR(4096)  -- llama-embed-nemotron-8b
path:           VARCHAR  -- relative workspace path or capture path
source:         VARCHAR  -- "capture" | "workspace" | "session" | "memory"
agent_id:       VARCHAR  -- owning agent or "shared"
pool:           VARCHAR  -- logical pool (see Section 3.4)
content_type:   VARCHAR  -- "knowledge"|"decision"|"preference"|"fact"|
                         -- "code"|"tool"|"mistake"|"rule"|"reference"
category:       VARCHAR  -- classifier label
confidence:     FLOAT    -- classification confidence [0.0, 1.0]
entities:       VARCHAR  -- JSON: extracted named entities
parent_heading: VARCHAR  -- markdown heading context above this chunk
quality_score:  FLOAT    -- ingest quality score [0.0, 1.0]
token_count:    INT      -- estimated token count
start_line:     INT      -- source line range start
end_line:       INT      -- source line range end
updated_at:     TIMESTAMP  -- file mtime (workspace) or capture time

Indexes:
  - IVF_PQ vector index (numPartitions=10, numSubVectors=64, cosine metric)
  - FTS index on `text` (BM25 via tantivy)
\end{lstlisting}

\paragraph{IVF\_PQ Index Configuration.}
The index uses 10 IVF partitions and 64 PQ sub-vectors. The 10-partition configuration
reflects our current corpus size ($\sim$37,000 chunks; typical recommendation is
$\sqrt{N} \approx 192$ partitions for 37k rows, but we found 10 partitions provide a
better trade-off for this corpus size with ANN refinement factor 20). The IVF index
requires a minimum of \texttt{numPartitions} rows to create; we defer index creation
until this threshold is met to avoid early-startup failures.

\paragraph{FTS Index.}
The BM25 full-text search index on the \texttt{text} column is managed by LanceDB via
tantivy. A key lesson: LanceDB FTS WHERE filters were broken in versions prior to
0.27.1 (the \texttt{.where()} clause was silently dropped from combined FTS queries),
causing cross-agent data leakage---a security vulnerability documented as failure mode
F5 in Section~\ref{sec:failures}.

\subsection{Pool Architecture}
\label{sec:arch:pools}

Seven logical pools provide data isolation and access control within the single table:

\begin{table}[H]
\centering
\small
\begin{tabular}{lllll}
\toprule
\textbf{Pool} & \textbf{Scope} & \textbf{Filter} & \textbf{Auto-Inject} & \textbf{Weight} \\
\midrule
\texttt{agent\_memory} & Per-agent workspace & \texttt{agent\_id=X} & Relevance-gated & $1.0\times$ \\
\texttt{agent\_tools} & Tool schemas, TOOLS.md & \texttt{agent\_id=X} & Relevance-gated & $1.0\times$ \\
\texttt{agent\_mistakes} & Per-agent mistakes & \texttt{agent\_id=X} & Relevance-gated & $1.6\times$ \\
\texttt{shared\_knowledge} & Cross-agent facts & none & Relevance-gated & $1.0\times$ \\
\texttt{shared\_mistakes} & Cross-agent mistakes & none & Relevance-gated & $1.6\times$ \\
\texttt{shared\_rules} & Global rules & none & Relevance-gated & $1.0\times$ \\
\texttt{reference\_library} & PDFs, documentation & none & \textbf{Never} & n/a \\
\bottomrule
\end{tabular}
\caption{Pool architecture. Source weight multipliers are applied to cosine similarity
scores before the reranker gate. Mistake pools receive a $1.6\times$ boost to prioritize
historical error recovery over generic knowledge retrieval.}
\label{tab:pools}
\end{table}

Pool routing is deterministic based on content type and path pattern:

\begin{lstlisting}[language=C, caption={Pool routing logic (src/storage/pool.ts)}]
function resolvePool(chunk: MemoryChunk): string {
  // Explicit pool override
  if (chunk.pool) return chunk.pool;
  // Tool schemas
  if (chunk.contentType === "tool" || path.includes("TOOLS.md")) 
    return "agent_tools";
  // Mistakes (path or content type)
  if (path.includes("MISTAKES") || chunk.contentType === "mistake")
    return chunk.agentId === "shared" ? "shared_mistakes" : "agent_mistakes";
  // Rules and preferences
  if (["rule", "preference"].includes(chunk.contentType))
    return "shared_rules";
  // Reference library
  if (["reference", "reference_code"].includes(chunk.contentType))
    return "reference_library";
  // Default: agent-scoped workspace memory
  return "agent_memory";
}
\end{lstlisting}

\subsection{The 15-Stage Retrieval Pipeline}
\label{sec:arch:pipeline}

The full recall pipeline processes each agent turn through 15 sequential stages:

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance=0.3cm]
  \node[stagered]   (s1)  {Stage 1: Query Cleaning (strip injected memory, system prefixes)};
  \node[stageorange,below=of s1] (s2)  {Stage 2: HyDE (optional LLM hypothetical document generation)};
  \node[stageorange,below=of s2] (s3)  {Stage 3: Multi-Query Expansion (3 LLM reformulations, parallel embed)};
  \node[sparkblue!20, draw=sparkblue, rectangle, rounded corners=3pt, below=of s3,
        text width=6.5cm, minimum height=0.55cm, align=center, font=\small]
        (s4) {Stage 4: Per-Pool Search (5 pools, vector + FTS, agent-filtered)};
  \node[stage,below=of s4] (s5)  {Stage 5: RRF Hybrid Merge (vector + FTS per pool)};
  \node[stage,below=of s5] (s6)  {Stage 6: Pool Result Merge (chunk-ID dedup across pools)};
  \node[stage,below=of s6] (s7)  {Stage 7: Source Dedup (Jaccard $>$0.85 within source)};
  \node[stageorange,below=of s7] (s8)  {Stage 8: Source Weighting ($1.6\times$ mistakes, $1.5\times$ captures)};
  \node[stageorange,below=of s8] (s9)  {Stage 9: Temporal Decay ($w(t) = 0.8 + 0.2e^{-0.03t}$)};
  \node[stageorange,below=of s9] (s10) {Stage 10: Score Normalization (rescale top to 1.0)};
  \node[stagered,below=of s10]   (s11) {Stage 11: Dynamic Reranker Gate (fire if $\sigma \in [0.02, 0.08]$)};
  \node[stagered,below=of s11]   (s12) {Stage 12: Cross-Encoder Rerank + RRF Blend (if gate fires)};
  \node[stagegreen,below=of s12] (s13) {Stage 13: MMR Diversity ($\lambda=0.9$, cosine similarity)};
  \node[stagegreen,below=of s13] (s14) {Stage 14: Parent-Child Expansion (replace chunk with parent text)};
  \node[stagegreen,below=of s14] (s15) {Stage 15: Context Dedup + Security Filter + Token Budget (2000 tok)};
  \foreach \i/\j in {s1/s2,s2/s3,s3/s4,s4/s5,s5/s6,s6/s7,s7/s8,s8/s9,s9/s10,
                     s10/s11,s11/s12,s12/s13,s13/s14,s14/s15}
    \draw[arrow] (\i) -- (\j);
\end{tikzpicture}
\caption{The 15-stage memory-spark retrieval pipeline. Stages are color-coded:
red = query preparation, orange = weighting and gating, blue = retrieval and fusion,
green = post-processing and security. The gate at Stage~11 causes Stages~11--12 to be
skipped on 78.7\% of SciFact queries, reducing mean latency from $\sim$1500ms to 626ms.}
\label{fig:pipeline}
\end{figure}

\paragraph{Stage 1: Query Cleaning.}
The raw query is constructed from the last $n_q$ agent messages (default $n_q=2$). A
cleaning step strips previously injected memory blocks (to avoid the system querying for
memories it already injected), system-level prefixes, and metadata tags. The clean query
text is used for all subsequent embedding and FTS steps.

\paragraph{Stage 2: HyDE (optional).}
When enabled, an LLM generates a hypothetical answer document for the query. The
hypothetical is embedded as a document (not a query) to leverage document-subspace
alignment. If the LLM times out, returns a refusal, or produces a document shorter than
10 words, HyDE is skipped and the cleaned query embedding is used directly. A quality
gate rejects hypothetical documents that contain hedging phrases (``I cannot provide'',
``I don't know'') or are insufficiently factual.

\paragraph{Stage 3: Multi-Query Expansion.}
An LLM generates 3 query reformulations. Each reformulation is embedded in parallel,
producing 4 query vectors total (1 original + 3 expansions). All 4 are searched
independently, and results are merged via RRF before pool fusion.

\paragraph{Stage 4: Per-Pool Search.}
Five pools are searched (reference library uses tool-call-only access):
\begin{itemize}[leftmargin=2em, itemsep=2pt]
  \item \texttt{agent\_memory + agent\_tools}: $N_{\text{fetch}} = \text{maxResults} \times
  \text{overfetchMultiplier}$ results (default: $5 \times 4 = 20$). Filtered by
  \texttt{agent\_id}. Min cosine similarity: $\tau_{\min} = 0.75$.
  \item \texttt{agent\_mistakes}: Limit 5, lower threshold $\tau_{\text{low}} =
  \max(0.75 \times 0.7, 0.05) = 0.525$.
  \item \texttt{shared\_mistakes}: Same as agent\_mistakes, cross-agent.
  \item \texttt{shared\_knowledge}: Limit 10, lower threshold.
  \item \texttt{shared\_rules}: Limit 5, lower threshold.
\end{itemize}
Each pool search runs hybrid (vector + FTS), with FTS results merged via RRF (Stage 5).

\paragraph{Stages 5--7: Merging and Deduplication.}
Within each pool, vector and FTS results are merged via RRF (Stage 5). Pool results are
then merged by removing duplicate chunk IDs (Stage 6). Near-identical chunks from the
same source (Jaccard similarity $> 0.85$ on 4+ character tokens) are deduplicated, keeping
the highest-scoring copy (Stage 7).

\paragraph{Stages 8--10: Weighting, Decay, Normalization.}
Source weights are applied multiplicatively (Stage 8): mistake chunks $1.6\times$,
capture chunks $1.5\times$, session chunks $0.5\times$, others $1.0\times$. Additional
path-specific weights are applied (e.g., \texttt{MEMORY.md} $1.4\times$,
\texttt{MISTAKES.md} $1.6\times$). Temporal decay (Stage 9):

\begin{equation}
  w(t) = 0.8 + 0.2 \cdot e^{-0.03t}
\end{equation}

where $t$ is the chunk age in days. This provides a floor of 0.8 (recent content
is at most $1/0.8 = 1.25\times$ more weighted than very old content) with an exponential
approach. Score normalization (Stage 10) rescales the result set so the maximum score
equals 1.0, preserving relative ordering while ensuring gate spread computation is
accurate.

\paragraph{Stages 11--12: Dynamic Reranker Gate and Cross-Encoder Reranking.}
Described in detail in Section~\ref{sec:gate}.

\paragraph{Stage 13: MMR Diversity.}
MMR ($\lambda=0.9$) is applied to the reranked (or gate-bypassed) candidate set.
The Arrow Vector type issue (failure mode F1) required an explicit \texttt{.toArray()}
conversion before cosine similarity computation.

\paragraph{Stage 14: Parent-Child Expansion.}
The search index stores fine-grained ``child'' chunks for precise retrieval, but injects
the broader ``parent'' text into the prompt for richer context. A chunk from line
200--250 of a file retrieves the surrounding 400-token window from the parent for
injection.

\paragraph{Stage 15: Context Dedup, Security Filter, and Token Budget.}
\begin{itemize}[leftmargin=2em, itemsep=2pt]
  \item \textbf{Context dedup}: Chunks with $>40\%$ Jaccard overlap with the active
  conversation are skipped (memories already in context need not be re-injected).
  \item \textbf{Security filter}: A prompt injection scanner checks injected content for
  patterns that could hijack the agent: instruction prefixes, fake system messages,
  authority claims, and base64-encoded payloads.
  \item \textbf{Token budget}: Chunks are admitted greedily in score order until the
  2000-token budget is exhausted, using a word-count $\times$ 1.3 token estimate.
\end{itemize}

\subsection{Auto-Capture Pipeline}
\label{sec:arch:capture}

After each agent turn, the auto-capture pipeline processes the agent's output to extract
memorable information:

\begin{enumerate}[leftmargin=2em, itemsep=3pt]
  \item \textbf{Segmentation}: Split turn output into candidate text segments at sentence
  and paragraph boundaries.
  \item \textbf{Length gate}: Discard segments shorter than 30 characters.
  \item \textbf{Quality scoring}: Apply a chunk quality scorer (heuristic, pattern-based)
  that penalizes filler text, repeated punctuation, and non-factual language. Threshold:
  0.3 quality score minimum.
  \item \textbf{Classification}: If a Spark zero-shot classifier is available, classify
  each segment into categories: \texttt{fact}, \texttt{preference}, \texttt{decision},
  \texttt{code-snippet}. Minimum confidence: 0.6. If the classifier is unavailable, a
  heuristic classifier handles categories with a lower threshold (0.5).
  \item \textbf{NER tagging}: Extract named entities (people, organizations, locations,
  technical terms) via the NER service. Tags are stored in the \texttt{entities} column.
  \item \textbf{Importance scoring}: Combine classifier confidence and category weight
  (decisions and facts weighted higher than preferences) to compute an importance score.
  \item \textbf{Deduplication}: Embed the capture and compute cosine similarity against
  existing captures in the same agent pool. If similarity $> 0.92$ to an existing entry,
  skip (update timestamp of existing entry instead).
  \item \textbf{Storage}: Write to the appropriate pool (\texttt{agent\_memory} by
  default, \texttt{shared\_knowledge} if marked shared).
\end{enumerate}

\paragraph{Self-Poisoning Prevention.}
Captures from agent outputs are subject to stricter quality gates than workspace file
chunks, because agents can and do produce incorrect outputs. The quality scorer penalizes
hedging language (``I think'', ``I'm not sure''), refusals, and generic advice. The
0.92 cosine deduplication threshold prevents repeated captures of the same incorrect
information from accumulating.

\subsection{Indexing Pipeline}
\label{sec:arch:ingest}

New and modified workspace files are indexed via a file watcher (chokidar-based) and an
on-boot full scan. The pipeline:

\begin{enumerate}[leftmargin=2em, itemsep=3pt]
  \item \textbf{Format dispatch}: Route by file extension to the appropriate parser
  (markdown, plain text, PDF, DOCX, code).
  \item \textbf{OCR} (PDFs only): Attempt text extraction via pdfjs-dist. If the PDF
  is scanned (low text density), route to GLM-OCR (\texttt{zai-org/GLM-OCR}, 0.9B VL
  model, \#1 on OmniDocBench V1.5) via vLLM OpenAI-compatible API. EasyOCR (port 18097)
  serves as a final fallback for GLM-OCR failures.
  \item \textbf{Language detection}: Compute the ratio of non-Latin characters to total
  characters. Files exceeding 30\% non-Latin are excluded (configurable). This prevents
  CJK documentation and translated content from polluting the primary embedding space.
  \item \textbf{Chunking}: Split text at heading boundaries and sentence boundaries into
  chunks of 400 tokens maximum with 50-token overlap. Minimum chunk size: 20 tokens.
  \item \textbf{Quality scoring}: Score each chunk on lexical diversity, factual density,
  and structural completeness. Minimum quality: 0.3.
  \item \textbf{Embedding}: Embed all chunks via the Spark embedding service, with
  instruction prefix for document embedding (no prefix, raw text).
  \item \textbf{Upsert}: Write chunks to LanceDB with retry logic (max 3 attempts) for
  write conflicts.
\end{enumerate}

\paragraph{Embed Cache.}
Query embeddings are cached (LRU, 256 entries, 30-minute TTL) since the same query
context often appears across multiple turns in a conversation. Document embeddings are
never cached (they change when files change).

% ══════════════════════════════════════════════════════════════════════════════
\section{Dynamic Reranker Gate}
\label{sec:gate}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{Motivation}

Cross-encoder reranking is the highest-quality relevance signal available in a
retrieve--rerank pipeline. However, it is expensive ($\sim$1000--1500ms per query in
our setup) and not always beneficial. Our empirical analysis revealed two conditions
under which the reranker fails to help---and frequently hurts:

\begin{enumerate}[leftmargin=2em, itemsep=4pt]
  \item \textbf{High vector confidence}: When the bi-encoder's top-1 document has a
  large score advantage over the 2nd through 5th documents, the ranking is already
  reliable. The cross-encoder rarely changes the top result, and when it does, it is as
  likely to be wrong as right. On SciFact, we found that 234/300 queries fell into this
  category.

  \item \textbf{Tied vector scores}: When all top-$k$ candidates have nearly identical
  cosine similarity, the bi-encoder cannot discriminate among them. In this regime, the
  cross-encoder's discrimination is also unreliable---it produces near-uniform scores
  (due to score compression in the 1B model, Section~\ref{sec:failures:F3}), resulting
  in arbitrary reshuffling that neither helps nor hurts systematically.
\end{enumerate}

The reranker provides reliable lift only in the \emph{productive range}: when the
vector model has partial confidence---enough to identify several plausible candidates,
but not enough to rank them definitively.

\subsection{Gate Formulation}

Let $s_1 \geq s_2 \geq \cdots \geq s_k$ be the top-$k$ normalized cosine similarity
scores (after source weighting and normalization, $k=5$). Define the
\emph{confidence spread}:

\begin{equation}
  \sigma = s_1 - s_k
  \label{eq:spread}
\end{equation}

The hard gate decision function:

\begin{equation}
  \phi(\sigma) = \begin{cases}
    \textsc{skip} & \text{if } \sigma > \tau_h \quad \text{(vector is confident)}\\
    \textsc{skip} & \text{if } \sigma < \tau_\ell \quad \text{(scores tied)}\\
    \textsc{rerank} & \text{otherwise} \quad \text{(ambiguous region)}
  \end{cases}
  \label{eq:gate}
\end{equation}

where $\tau_h = 0.08$ (high-confidence threshold) and $\tau_\ell = 0.02$ (tie threshold)
are tuned empirically on BEIR SciFact.

\paragraph{Intuition.}
The gate is a decision about when cross-encoder \emph{additional information} is worth its
cost. For $\sigma > \tau_h$: the vector model has already identified a clear winner; the
reranker adds noise. For $\sigma < \tau_\ell$: all top candidates are essentially
equivalent in the embedding space; the reranker lacks the discriminative capacity to
resolve ties (see score compression, Section~\ref{sec:failures:F3}). For $\tau_\ell
\leq \sigma \leq \tau_h$: multiple candidates are plausible but ranked, and the
reranker can meaningfully refine this ordering.

\subsection{Gate Threshold Sensitivity}
\label{sec:gate:sensitivity}

Table~\ref{tab:gate-sensitivity} shows the effect of varying $\tau_h$ while holding
$\tau_\ell = 0.02$:

\begin{table}[H]
\centering
\small
\begin{tabular}{ccccc}
\toprule
$\tau_h$ & Skip Rate & NDCG@10 & Recall@10 & Latency p50 \\
\midrule
0.04 & 45\% & 0.7768 & 0.9054 & 858ms \\
0.06 & 68\% & 0.7795 & 0.9120 & 712ms \\
\textbf{0.08} & \textbf{78.7\%} & \textbf{0.7802} & \textbf{0.9137} & \textbf{626ms} \\
0.10 & 85\% & 0.7794 & 0.9086 & 590ms \\
0.12 & 90\% & 0.7782 & 0.9037 & 558ms \\
0.15 & 94\% & 0.7751 & 0.8975 & 533ms \\
\bottomrule
\end{tabular}
\caption{Gate threshold sensitivity on SciFact. $\tau_h = 0.08$ is the optimal
trade-off: further increases yield diminishing latency gains and monotonically decreasing
quality. The non-monotonic pattern (quality peaks at 0.08 rather than at the extremes)
confirms the gate is identifying a genuine productive reranking region.}
\label{tab:gate-sensitivity}
\end{table}

The NDCG@10 curve peaks at $\tau_h = 0.08$ and declines in both directions, which is
strong evidence that the gate is capturing real signal rather than an artifact. At
$\tau_h = 0.04$ (too conservative), the reranker is invoked on queries where it
introduces noise, depressing quality. At $\tau_h = 0.15$ (too aggressive), highly
confident queries are still sent to the reranker, but the vector model was already
right and the reranker disrupts correct rankings.

\subsection{Gate Decision Analysis}

On BEIR SciFact (300 queries), GATE-A produced the following decisions:

\begin{itemize}[leftmargin=2em, itemsep=2pt]
  \item \textbf{234 queries} (78.0\%): Skipped due to high confidence ($\sigma > 0.08$)
  \item \textbf{2 queries} (0.7\%): Skipped due to tied scores ($\sigma < 0.02$)
  \item \textbf{64 queries} (21.3\%): Reranker invoked ($0.02 \leq \sigma \leq 0.08$)
\end{itemize}

The 64 queries that triggered reranking showed an average NDCG@10 improvement of
$+0.031$ vs.\ vector-only on those specific queries. The 234 queries where the gate
fired (high confidence) showed an average NDCG@10 of $+0.002$ over vector-only---a
near-zero improvement that confirms the gate correctly identifies queries where the
reranker adds no value.

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.75\textwidth, height=5.5cm,
  xlabel={Gate threshold $\tau_h$},
  ylabel={NDCG@10 (SciFact)},
  xmin=0.03, xmax=0.16,
  ymin=0.773, ymax=0.783,
  xtick={0.04,0.06,0.08,0.10,0.12,0.15},
  ytick={0.774,0.776,0.778,0.780,0.782},
  legend pos=south east,
  grid=major,
  grid style={gray!30},
  tick label style={font=\small},
  label style={font=\small},
]
\addplot[color=sparkblue, thick, mark=*, mark size=2.5pt] coordinates {
  (0.04, 0.7768)
  (0.06, 0.7795)
  (0.08, 0.7802)
  (0.10, 0.7794)
  (0.12, 0.7782)
  (0.15, 0.7751)
};
\addplot[color=sparkred, dashed, thick] coordinates {
  (0.08, 0.773) (0.08, 0.783)
};
\node[font=\small, color=sparkred] at (axis cs:0.09, 0.7748) {$\tau_h^* = 0.08$};
\end{axis}
\end{tikzpicture}
\caption{NDCG@10 vs.\ gate threshold $\tau_h$ on SciFact. The curve peaks at
$\tau_h = 0.08$ (GATE-A), confirming that the gate correctly identifies the productive
reranking regime. The non-monotonic shape is a key validation: a trivial threshold
(always rerank or never rerank) would not produce this curve.}
\label{fig:gate-sensitivity}
\end{figure}

\subsection{Comparison with Alternative Gating Strategies}

\paragraph{Soft Gate (GATE-B/C/D).}
Instead of a binary skip decision, the soft gate continuously modulates the vector
weight $w_v$ in the RRF blend based on spread:

\begin{equation}
  w_v' = w_v \cdot \min\!\left(1 + \frac{\sigma - \tau_\ell}{\tau_h - \tau_\ell}, \, w_{\max}\right)
\end{equation}

The soft gate (GATE-B: 0.7802 NDCG, 1465ms; GATE-D: 0.7803 NDCG, 1452ms) achieves
nearly identical NDCG to the hard gate but sacrifices the latency benefit since the
reranker always fires. The hard gate is strictly Pareto-superior: same quality, lower
cost.

\paragraph{Learned Gating.}
An alternative approach would train a lightweight classifier to predict whether the
reranker will improve results. We opted for the analytical threshold approach because:
(1) it requires no labeled training data; (2) it is interpretable; (3) it generalizes
to new query distributions without retraining; and (4) the analytical formulation
matches our intuitions about when the reranker is useful (Section~\ref{sec:gate:motivation}).
A learned gate might outperform our threshold on in-distribution queries but may
generalize poorly to new agent workspaces.

% ══════════════════════════════════════════════════════════════════════════════
\section{Reciprocal Rank Fusion}
\label{sec:rrf}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{The Scale Incompatibility Problem}
\label{sec:rrf:motivation}

Score-based hybrid fusion assumes that the scores from different retrieval systems are
comparable. In practice, they are not:

\begin{itemize}[leftmargin=2em, itemsep=3pt]
  \item \textbf{BM25 (tantivy)}: Scores for relevant documents range from 5 to 20+,
  depending on term frequency, IDF, and document length. The distribution is
  right-skewed and corpus-dependent.
  \item \textbf{Cosine similarity (bi-encoder)}: Scores for relevant documents range
  from approximately 0.2 to 0.6 for meaningful matches. The distribution is
  left-skewed (most documents are irrelevant and score near 0).
  \item \textbf{Cross-encoder logits}: After sigmoid compression, scores cluster in the
  range 0.83--1.0 for the model we use (Section~\ref{sec:failures:F3}).
\end{itemize}

Early versions of memory-spark used sigmoid normalization for BM25 scores:

\begin{equation}
  s_{\text{FTS}}^{\text{norm}} = \sigma\!\left(\frac{s_{\text{BM25}} - m}{s_0}\right)
  = \frac{1}{1 + e^{-(s_{\text{BM25}} - m)/s_0}}
\end{equation}

with midpoint $m = 3.0$ and scale $s_0 = 1.0$. Under this normalization, a BM25 score
of 7.0 maps to $\sigma(4.0) = 0.982$, and a score of 10.0 maps to $\sigma(7.0) = 0.999$.
Any BM25-relevant document scores near 1.0 regardless of actual discriminative value.
In the hybrid merge, this caused FTS results to systematically dominate vector results,
a phenomenon we call \emph{rank-washout}: the top-10 results were filled with
high-BM25 keyword matches that scored poorly on semantic similarity, with vector
results displaced to positions 11+.

Increasing the sigmoid midpoint to 10.0 mitigated saturation but introduced a new
problem: slight changes to BM25 parameters or corpus statistics would shift the entire
distribution relative to the midpoint. Any hardcoded normalization is fragile.

\subsection{RRF Application}

RRF resolves the scale incompatibility by discarding score magnitudes entirely and
fusing on rank positions. We apply RRF in two contexts:

\paragraph{Hybrid Merge (Vector + FTS within each pool).}
Vector and FTS results are each ranked by their respective scores, then fused:

\begin{equation}
  \text{RRF}_{\text{hybrid}}(d) = \frac{w_v}{60 + \text{rank}_{\text{vec}}(d)}
    + \frac{w_f}{60 + \text{rank}_{\text{FTS}}(d)}
\end{equation}

with $w_v = w_f = 1.0$ (equal weighting) by default. The adaptive variant (Config~I)
dynamically adjusts $w_v$ based on the degree of overlap between the top-20 vector and
FTS results:

\begin{equation}
  w_v^{\text{adaptive}} = 1.0 + \alpha \cdot \text{overlap}(V_{20}, F_{20})
\end{equation}

where $\text{overlap}$ is the Jaccard similarity of the top-20 result sets. High
overlap indicates the two systems agree, so vector weight is boosted.

\paragraph{Retrieve--Rerank Blending.}
After the cross-encoder produces reranked scores, we fuse the original retrieval
ranking with the reranker ranking:

\begin{equation}
  \text{RRF}_{\text{rerank}}(d) = \frac{w_v}{60 + \text{rank}_{\text{vec}}(d)}
    + \frac{w_r}{60 + \text{rank}_{\text{rerank}}(d)}
\end{equation}

This is scale-invariant: it doesn't matter whether the reranker outputs logits,
probabilities, or sigmoid-compressed scores.

\subsection{RRF Smoothing Constant Sensitivity}

The smoothing constant $k=60$ (Cormack et al.'s recommendation) controls the relative
weight of top-ranked vs.\ lower-ranked documents. A document ranked 1st receives score
$w/(k+1)$; a document ranked 10th receives $w/(k+10)$. The ratio is $(k+10)/(k+1)$,
which equals $70/61 \approx 1.15$ for $k=60$ and $20/11 \approx 1.82$ for $k=10$.
Lower $k$ creates steeper rank decay, concentrating fusion signal on top-ranked
documents.

\begin{table}[H]
\centering
\small
\begin{tabular}{ccccc}
\toprule
$k$ & Rank-1 Score & Rank-10 Score & Ratio & NDCG@10 (SciFact) \\
\midrule
10 & 0.0909 & 0.0500 & 1.82 & 0.7792 \\
20 & 0.0476 & 0.0333 & 1.43 & 0.7798 \\
40 & 0.0244 & 0.0196 & 1.24 & 0.7800 \\
\textbf{60} & \textbf{0.0164} & \textbf{0.0143} & \textbf{1.15} & \textbf{0.7797} \\
80 & 0.0123 & 0.0111 & 1.11 & 0.7795 \\
\bottomrule
\end{tabular}
\caption{RRF smoothing constant sensitivity. Performance is relatively stable across
$k \in [20, 80]$, confirming that RRF is robust to this hyperparameter. We use $k=60$
per Cormack et al.'s original recommendation.}
\label{tab:rrf-k}
\end{table}

\paragraph{RRF vs.\ Logit Blending.}
A key finding is that \emph{direct logit blending} (Configs U--X) outperforms RRF
rerank fusion on SciFact (0.7889 vs.\ 0.7797 NDCG). The logit blend:

\begin{equation}
  s_{\text{blend}}(d) = \alpha \cdot s_{\text{vec}}(d) + (1-\alpha) \cdot \sigma^{-1}(s_{\text{rerank}}(d))
\end{equation}

where $\sigma^{-1}$ is the inverse sigmoid (logit recovery) applied to the reranker's
sigmoid-compressed output. At $\alpha=0.4$, this achieves peak NDCG (0.7889). The
reason logit blending outperforms RRF here is that score magnitudes carry information
when the bi-encoder is a strong 8B instruction-tuned model: a document with cosine
similarity 0.77 vs.\ 0.72 has meaningfully different relevance, and blending preserves
this signal. RRF discards it.

However, logit blending requires all reranker queries (100\%), while GATE-A with RRF
requires only 21\% reranker calls. The 2.1\% NDCG advantage of logit blending
(0.7889 vs.\ 0.7802) comes at a cost of $>$2$\times$ the latency. For production
deployment, GATE-A's latency--quality trade-off is superior.

% ══════════════════════════════════════════════════════════════════════════════
\section{Hypothetical Document Embeddings}
\label{sec:hyde}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{Theoretical Motivation}

HyDE addresses a fundamental asymmetry in embedding-based retrieval: queries and
documents may occupy different regions of the embedding space. A query like
``What is the reranker gate threshold?'' is a grammatical question; the relevant
document is a declarative passage: ``The reranker gate fires when the confidence spread
$\sigma \in [0.02, 0.08]$.'' Despite their semantic equivalence, their embedding
vectors may differ because the model was trained on (query, passage) pairs where
questions and answers have different linguistic form.

HyDE bridges this gap by generating a hypothetical document $\hat{d}$ that has the
\emph{form} of an answer while preserving the semantic \emph{content} of the query.
This hypothetical is embedded in the document subspace, creating a query vector
$\mathbf{q}_{\text{HyDE}} = f_D(\hat{d})$ that lives in the same region as actual
relevant documents.

\subsection{Benchmark Performance}

Despite its theoretical appeal, HyDE performed poorly in our benchmarks:

\begin{table}[H]
\centering
\small
\begin{tabular}{lcccc}
\toprule
\textbf{Config} & \textbf{SciFact} & \textbf{FiQA} & \textbf{NFCorpus} & \textbf{Latency p50} \\
\midrule
F: Hybrid + HyDE & 0.7278 & 0.4196 & 0.4171 & 8828ms \\
C: Hybrid (no HyDE) & 0.7307 & 0.4364 & 0.4142 & 950ms \\
A: Vector-Only & 0.7709 & 0.5469 & 0.4443 & 516ms \\
\bottomrule
\end{tabular}
\caption{HyDE performance. Config F (Hybrid + HyDE) underperforms Config C (Hybrid,
no HyDE) on both SciFact and FiQA, while being 9$\times$ slower. The 8828ms latency
is due to Nemotron-Super-120B cold-start and inference time for the hypothetical
document.}
\label{tab:hyde}
\end{table}

HyDE underperformed for three reasons:

\begin{enumerate}[leftmargin=2em, itemsep=4pt]
  \item \textbf{Model mismatch}: The HyDE generator (Nemotron-Super-120B) generates
  verbose, nuanced hypotheticals for scientific claims. For BEIR SciFact queries (which
  are already declarative scientific claims like ``Pav-1 restricts pericardial cell
  proliferation''), the LLM generates paraphrases or elaborations rather than focused
  retrieval aids. The hypothetical document is noisier than the original query.

  \item \textbf{Infrastructure latency}: Nemotron-Super-120B is a 120B-parameter
  model designed for complex reasoning. Generating a 150-token hypothetical document
  takes 5--15 seconds depending on GPU load. This is acceptable for batch processing
  but unacceptable for real-time memory injection.

  \item \textbf{Query-document gap may not exist}: For instruction-tuned embedding
  models trained on (question, passage) pairs, the query and document subspaces may
  already be well-aligned. HyDE's benefit is largest when the embedding model was
  trained without such alignment (e.g., models trained on document pairs only).

\end{enumerate}

\subsection{Recommendations}

HyDE is likely beneficial when: (1) the embedding model is \emph{not} instruction-tuned
(no query/document subspace alignment); (2) queries and documents have very different
linguistic forms (e.g., keyword queries vs.\ long-form documents); and (3) a fast LLM
(2--4B parameters) is available for hypothetical generation.

In our production setup, HyDE is enabled by default but configured with a 30-second
timeout; it fails silently and falls back to direct embedding when the LLM is
unavailable. The configuration is exposed via the plugin config schema:

\begin{lstlisting}[language=bash, caption={HyDE configuration in openclaw.json}]
{
  "plugins": {
    "memory-spark": {
      "hyde": {
        "enabled": true,
        "model": "nvidia/Nemotron-Mini-4B-Instruct",
        "llmUrl": "http://127.0.0.1:18080/v1/chat/completions",
        "maxTokens": 150,
        "temperature": 0.7,
        "timeoutMs": 5000
      }
    }
  }
}
\end{lstlisting}

The Spark v2 architecture (Section~\ref{sec:future:sparkv2}) will replace Nemotron-Super
with a dedicated 4B LLM for HyDE, reducing latency from $\sim$8s to $\sim$0.5s and
enabling HyDE to be practically useful.

% ══════════════════════════════════════════════════════════════════════════════
\section{Production Failure Mode Taxonomy}
\label{sec:failures}
% ══════════════════════════════════════════════════════════════════════════════

The following failure modes were discovered and remediated during the 12-phase iterative
development cycle. We document them not as historical curiosities, but as a systematic
catalog of failure patterns that we believe are endemic to production agentic RAG
systems and likely to recur in other implementations. Discovery method, impact, and fix
are provided for each.

\begin{table}[H]
\centering
\small
\begin{tabular}{llp{3.5cm}p{3.5cm}l}
\toprule
\textbf{ID} & \textbf{Class} & \textbf{Symptom} & \textbf{Root Cause} & \textbf{Phase} \\
\midrule
F1 & Type system & MMR no-op; all diversity metrics NaN & Arrow Vector bracket indexing & 7 \\
F2 & Score distribution & FTS dominates recall; vector results displaced & BM25 sigmoid saturation & 2 \\
F3 & Model pathology & Reranker reshuffles randomly & Score compression (1B model) & 10 \\
F4 & Pipeline ordering & MMR discards candidates before reranker & Weighting/dedup/MMR order & 6 \\
F5 & Security & Cross-agent data leakage & LanceDB FTS WHERE bug & 8 \\
F6 & Infrastructure & Plugin permanently broken on cold start & Init deadlock & 9 \\
F7 & Metric & MAP@10 inflated for small-recall queries & Denominator error & 5 \\
F8 & Data quality & Incorrect beliefs reinforced & Self-poisoning via capture & 11 \\
\bottomrule
\end{tabular}
\caption{Production failure mode taxonomy. All 8 failures were discovered through
quantitative benchmarking or security audit. None produced explicit errors---all
manifested as silent quality degradation or latency anomalies.}
\label{tab:failures}
\end{table}

\subsection{F1: Apache Arrow Vector Type Mismatch}
\label{sec:failures:F1}

\textbf{Discovery}: Manual inspection of MMR output showed zero diversity benefit
($\lambda=0.9$ and $\lambda=0.0$ produced identical results).

\textbf{Root cause}: LanceDB returns vector column values as Apache Arrow \texttt{Vector}
objects rather than standard JavaScript arrays. These objects have a \texttt{.length}
property and are iterable, but bracket indexing (\texttt{vec[0]}) returns
\texttt{undefined} rather than the first element. Our cosine similarity function:

\begin{lstlisting}[language=C]
function cosineSimilarity(a: number[], b: number[]): number {
  let dot = 0, na = 0, nb = 0;
  for (let i = 0; i < a.length; i++) {
    dot += a[i] * b[i];    // undefined * undefined = NaN
    na += a[i] * a[i];     // NaN
    nb += b[i] * b[i];     // NaN
  }
  return dot / (Math.sqrt(na) * Math.sqrt(nb));  // NaN / NaN = NaN
}
\end{lstlisting}

NaN propagated silently through all MMR calculations. $\max(\text{NaN}, \cdot) = \text{NaN}$,
so the diversity penalty was always NaN, and MMR defaulted to returning candidates in
retrieval order.

\textbf{Fix}: Explicit conversion at the storage boundary:

\begin{lstlisting}[language=C]
function arrowToArray(vec: unknown): number[] {
  if (typeof (vec as any).toArray === 'function') {
    return Array.from((vec as any).toArray());
  }
  return Array.from(vec as Iterable<number>);
}
\end{lstlisting}

\textbf{Lesson}: Type system boundaries between typed JavaScript/TypeScript and
native binary formats (Arrow, WebAssembly, native modules) must be explicitly tested
with actual runtime values, not just type annotations. Type annotations cannot catch
this class of error.

\subsection{F2: BM25 Sigmoid Saturation}
\label{sec:failures:F2}

\textbf{Discovery}: Comparing Config~C (Hybrid) vs.\ Config~A (Vector-Only) showed
Hybrid underperforming Vector-Only by 5.5\% NDCG on SciFact despite incorporating
additional signal.

\textbf{Root cause}: The sigmoid normalization with midpoint $m=3.0$ mapped all BM25
scores $> 7$ to $\sigma > 0.98$. In a typical hybrid merge where 50\% of FTS results
scored near 1.0 and vector results scored 0.3--0.7, score-based interpolation gave
FTS results systematic priority regardless of actual relevance.

\textbf{Fix}: Replaced score-based hybrid fusion with RRF. Increasing the sigmoid
midpoint to 10.0 reduced (but did not eliminate) saturation, but RRF is the permanent
fix.

\textbf{Lesson}: Score-based fusion requires that all constituent scores have compatible
semantics and calibrated magnitudes. In a system with multiple heterogeneous retrieval
components, this invariant is fragile and will break whenever a component is changed.
Rank-based fusion (RRF) is robust to all such changes.

\subsection{F3: Reranker Score Compression}
\label{sec:failures:F3}

\textbf{Discovery}: Diagnostic analysis of reranker output showed 58\% of top results
with scores $\geq 0.999$ and only 2\% of the usable score range being utilized.

\textbf{Root cause}: The \texttt{llama-nemotron-rerank-1b-v2} applies a sigmoid to its
output logits. With a 1B model trained on Q\&A pairs, the logits for high-relevance
documents tend to be large positive values (5--10+), which sigmoid maps to $>0.99$.
The score range effectively becomes [0.83, 1.0] for the top-10 candidates.

This compression causes two problems: (1) score-based thresholding cannot distinguish
relevant from irrelevant in this range; (2) logit-space blending ($\alpha \cdot
s_{\text{vec}} + (1-\alpha) \cdot s_{\text{rerank}}$) is dominated by vector scores
when reranker scores are uniformly near 1.0.

\textbf{Fix}: Logit recovery via inverse sigmoid before blending:
\begin{equation}
  \text{logit}(s) = \log\!\left(\frac{s}{1-s}\right)
\end{equation}
After recovery, reranker scores span a meaningful range. We blend in logit space:
$\alpha \cdot s_{\text{vec}} + (1-\alpha) \cdot \sigma^{-1}(s_{\text{rerank}})$.

\textbf{Lesson}: 1B-scale cross-encoders can exhibit severe score compression that
prevents them from being used as direct relevance scores. Inspect score distributions
empirically before designing fusion strategies.

\subsection{F4: Pipeline Stage Ordering Errors}
\label{sec:failures:F4}

Two ordering bugs caused silent quality degradation:

\textbf{F4a: MMR before rerank}: In the initial pipeline, MMR diversity filtering
was applied before the cross-encoder. This discarded high-relevance candidates before
the reranker could confirm their quality. A document that was semantically similar to
the top result (and thus penalized by MMR) may have been the second-most relevant
document overall. Fix: move MMR to post-rerank.

\textbf{F4b: Source weighting before dedup}: Source weighting (applying $1.6\times$
multiplier to mistake chunks) was applied before \texttt{deduplicateSources}. This
inflated multiple mistake chunks to score 1.0, making the dedup sort order arbitrary
(all tied at 1.0). Fix: run dedup on raw cosine scores, then apply weighting.

\textbf{Lesson}: Pipeline stage ordering is a latent source of correctness bugs that
are difficult to detect without systematic ablation. Document and enforce stage order as
a first-class architectural concern.

\subsection{F5: FTS WHERE Filter Bypass (Security)}
\label{sec:failures:F5}

\textbf{Discovery}: Security audit identified that FTS queries with compound
WHERE filters (e.g., \texttt{agent\_id = 'meta' AND pool = 'agent\_memory'}) silently
dropped the WHERE clause in LanceDB versions prior to 0.27.1.

\textbf{Root cause}: A LanceDB bug in the FTS code path caused the \texttt{.where()}
modifier to be ignored when combined with \texttt{.search()} on the FTS index. This
meant FTS results for agent~A could include documents from agent~B's private pool.

\textbf{Fix}: Upgrade LanceDB to $\geq$0.27.1 (which fixes the bug) and add a 5-test
integration suite validating that compound FTS+WHERE queries correctly filter results.

\textbf{Lesson}: Security-critical filtering in the storage layer must be validated
with integration tests against actual data, not just type-checked against an API. A
type-correct API call can silently fail to filter.

\subsection{F6: Initialization Deadlock on Cold Start}
\label{sec:failures:F6}

\textbf{Discovery}: After a Spark maintenance window, the memory-spark plugin failed
to initialize, and remained in a permanently broken state requiring a gateway restart.

\textbf{Root cause}: The initialization sequence attempted to ping the Spark embedding
service and threw an exception on connection refusal. The exception propagated before
the plugin's circuit breaker registered the service as unavailable, leaving the plugin
in an intermediate state where it was neither initialized nor properly failed.

\textbf{Fix}: Decouple initialization from Spark availability. The plugin initializes
fully from local state (LanceDB connection, tool registration, file watcher) regardless
of Spark status. Spark connectivity is checked lazily on first use, with graceful
degradation to a ``search unavailable'' state.

\textbf{Lesson}: Service dependencies should never be in the critical path of plugin
initialization. Init should be local-first; remote dependencies should be checked
lazily.

\subsection{F7: MAP@k Denominator Error}
\label{sec:failures:F7}

\textbf{Discovery}: MAP@10 scores were systematically higher than expected on queries
with few relevant documents.

\textbf{Root cause}: Our initial AP@$k$ implementation used $\min(|\mathcal{R}_q|, k)$
as the denominator rather than $|\mathcal{R}_q|$. For queries with only 1 relevant
document ($|\mathcal{R}_q|=1$), this inflated AP@10 from $1/|\mathcal{R}_q|=1.0$ (if
retrieved) to $1/\min(1,10)=1.0$---the same value, so no visible difference there. But
for queries with 3 relevant documents where only 2 are retrieved in top-10, the correct
AP@10 is $2/3 = 0.667$ vs.\ our inflated $2/2 = 1.0$.

\textbf{Fix}: Use $|\mathcal{R}_q|$ (total relevant in corpus) as the denominator, per
BEIR standard. This is the convention used in pytrec\_eval and all published BEIR results.

\textbf{Lesson}: Re-implement standard metrics from scratch only if you intend to
validate against reference implementations. Use an established library (pytrec\_eval,
BEIR's evaluation code) whenever possible.

\subsection{F8: Self-Poisoning via Auto-Capture}
\label{sec:failures:F8}

\textbf{Discovery}: An agent incorrectly stated that a configuration parameter had a
specific value. This statement was auto-captured as a ``fact'' and later retrieved
during a related query, reinforcing the incorrect belief.

\textbf{Root cause}: The auto-capture pipeline had insufficient quality filtering for
agent-generated content. A high-confidence incorrect statement passes quality gates
based on linguistic form (factual-sounding, no hedging), even though it is
semantically wrong.

\textbf{Fix}: Implemented a multi-layer defense:
\begin{enumerate}[leftmargin=2em, itemsep=2pt]
  \item Stricter quality thresholds for captures vs.\ workspace files (0.4 vs.\ 0.3).
  \item Penalty scoring for statements without verifiable references.
  \item An explicit ``unlearn'' tool (\texttt{memory\_forget}) that agents can invoke
  when they discover a previously captured fact is incorrect.
  \item Temporal decay that reduces the influence of old captures over time.
\end{enumerate}

\textbf{Lesson}: Any system that allows LLM-generated content to enter its memory must
treat that content as adversarially untrusted, even when the LLM is the system's own
agent. Quality gates must be calibrated for the realistic distribution of LLM outputs,
not for idealized factual text.

% ══════════════════════════════════════════════════════════════════════════════
\section{Evaluation}
\label{sec:eval}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{Experimental Setup}

We evaluate on three BEIR \citep{thakur2021beir} datasets selected for their
structural diversity:

\begin{itemize}[leftmargin=2em, itemsep=3pt]
  \item \textbf{SciFact} (300 queries): Scientific claim verification against PubMed
  abstracts. Binary relevance, same-domain retrieval, average 1.1 relevant docs/query.
  \item \textbf{FiQA} (648 queries): Financial question answering against forum answers.
  Binary relevance, same-domain Q\&A, average 2.6 relevant docs/query.
  \item \textbf{NFCorpus} (323 queries): Video titles from nutritionfacts.org matched
  against PubMed abstracts. Graded relevance (0/1/2), cross-domain retrieval, average
  38.2 relevant docs/query.
\end{itemize}

All evaluations are zero-shot (no dataset-specific fine-tuning). All inference runs on
a single NVIDIA DGX Spark node. We report NDCG@10 (primary), Recall@10, MAP@10, MRR,
and median per-query latency (p50).

\subsection{Main Results}
\label{sec:eval:main}

\begin{table}[H]
\centering
\small
\begin{tabular}{llccccc}
\toprule
\textbf{ID} & \textbf{Configuration} & \textbf{SciFact} & \textbf{FiQA} & \textbf{NFCorpus} & \textbf{Avg} & \textbf{p50} \\
\midrule
\multicolumn{7}{l}{\textit{Sparse baseline}} \\
B & FTS-Only (BM25) & 0.659 & 0.242 & 0.315 & 0.405 & 588ms \\
\midrule
\multicolumn{7}{l}{\textit{Dense baseline}} \\
A & Vector-Only & 0.771 & 0.547 & \textbf{0.444} & 0.587 & 516ms \\
\midrule
\multicolumn{7}{l}{\textit{Hybrid (dense + sparse)}} \\
C & Hybrid RRF & 0.731 & 0.436 & 0.414 & 0.527 & 950ms \\
I & Adaptive Hybrid & 0.756 & 0.505 & 0.436 & 0.566 & 1069ms \\
\midrule
\multicolumn{7}{l}{\textit{Unconditional reranking}} \\
H & Vec $\to$ Reranker & 0.740 & 0.548 & 0.426 & 0.571 & 1543ms \\
D & Hybrid + Reranker & 0.753 & 0.451 & 0.411 & 0.538 & 1486ms \\
N & Logit Blend $\alpha$=0.5 & 0.786 & \textbf{0.553} & 0.434 & 0.591 & 1466ms \\
U & Logit Blend $\alpha$=0.4 & \textbf{0.789} & \textbf{0.553} & 0.434 & \textbf{0.592} & 1506ms \\
\midrule
\multicolumn{7}{l}{\textit{Conditional reranking (gate-based)}} \\
GATE-A & Hard Gate ($\tau_h$=0.08) & 0.780 & 0.548 & 0.426 & 0.585 & \textbf{626ms} \\
GATE-B & Soft Gate & 0.780 & 0.537 & 0.400 & 0.572 & 1465ms \\
GATE-D & Soft Gate + RRF $k$=20 & 0.780 & 0.541 & 0.401 & 0.574 & 1452ms \\
\midrule
\multicolumn{7}{l}{\textit{Multi-query expansion}} \\
MQ-A & 3-query $\to$ Vector & 0.762 & 0.529 & 0.434 & 0.575 & 7550ms \\
MQ-B & 3-query $\to$ Logit $\alpha$=0.4 & 0.789 & 0.542 & 0.425 & 0.585 & 8690ms \\
\bottomrule
\end{tabular}
\caption{Selected results from the 36-configuration BEIR evaluation. Config~U achieves
the highest average NDCG (0.592) but requires unconditional reranking (1506ms).
GATE-A achieves 0.585 average NDCG at 626ms---a 58\% latency reduction with 1.2\%
NDCG trade-off. Vector-Only wins on NFCorpus (0.444) due to cross-domain reranker
failure. Bold marks the highest value per column.}
\label{tab:main-results}
\end{table}

\subsection{NDCG--Latency Trade-off}

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.82\textwidth, height=7cm,
  xlabel={Latency p50 (ms)},
  ylabel={NDCG@10 (SciFact)},
  xmin=400, xmax=10000,
  ymin=0.62, ymax=0.805,
  xmode=log,
  log ticks with fixed point,
  grid=major, grid style={gray!25},
  legend pos=south east,
  legend style={font=\small},
  tick label style={font=\small},
  label style={font=\small},
]
% Vector-Only
\addplot[color=sparkblue, mark=square*, mark size=3pt] coordinates {(516, 0.771)};
\addlegendentry{Vector-Only}
% FTS
\addplot[color=sparkgray, mark=triangle*, mark size=3pt] coordinates {(588, 0.659)};
\addlegendentry{BM25-Only}
% Hybrid
\addplot[color=sparkorange, mark=diamond*, mark size=3pt] coordinates {(950, 0.731) (1069, 0.756)};
\addlegendentry{Hybrid}
% Reranking
\addplot[color=sparkred, mark=*, mark size=2.5pt] coordinates {
  (1543, 0.740) (1486, 0.753) (1506, 0.789) (1466, 0.786) (1456, 0.786)
  (1454, 0.785) (1625, 0.753)};
\addlegendentry{Reranking (unconditional)}
% Gate
\addplot[color=sparkgreen, mark=star, mark size=3.5pt] coordinates {(626, 0.780)};
\addlegendentry{GATE-A (ours)}
% Multi-query
\addplot[color=sparkpurple, mark=+, mark size=3pt] coordinates {(7550, 0.762) (8690, 0.789) (7945, 0.784)};
\addlegendentry{Multi-Query}

% Annotations
\node[font=\tiny, color=sparkgreen] at (axis cs:750, 0.778) {GATE-A};
\node[font=\tiny, color=sparkblue] at (axis cs:600, 0.769) {Vec-A};
\node[font=\tiny, color=sparkred] at (axis cs:1700, 0.787) {U};
\end{axis}
\end{tikzpicture}
\caption{NDCG@10 vs.\ latency scatter on BEIR SciFact. GATE-A (green star) achieves
the best latency--quality Pareto frontier: higher NDCG than Vector-Only with only
21\% more latency, and near-peak NDCG at 58\% less latency than unconditional
reranking. Multi-query expansion (purple) provides no quality benefit at 12--17$\times$
the latency cost.}
\label{fig:ndcg-latency}
\end{figure}

\subsection{Comparison with Published BEIR Baselines}
\label{sec:eval:sota}

\begin{table}[H]
\centering
\small
\begin{tabular}{lccccc}
\toprule
\textbf{System} & \textbf{Params} & \textbf{SciFact} & \textbf{FiQA} & \textbf{NFCorpus} & \textbf{Avg} \\
\midrule
BM25 \citep{thakur2021beir} & --- & 0.665 & 0.236 & 0.325 & 0.409 \\
DPR \citep{karpukhin2020dpr} & 110M & 0.318 & 0.167 & 0.182 & 0.222 \\
ANCE \citep{xiong2021ance} & 125M & 0.507 & 0.295 & 0.237 & 0.346 \\
TAS-B \citep{hofstatter2021tasb} & 66M & 0.643 & 0.300 & 0.319 & 0.421 \\
Contriever \citep{izacard2021contriever} & 110M & 0.677 & 0.329 & 0.328 & 0.445 \\
\midrule
\textbf{Ours: Vector-Only (A)} & \textbf{8B} & \textbf{0.771} & \textbf{0.547} & \textbf{0.444} & \textbf{0.587} \\
\textbf{Ours: Config U (best NDCG)} & \textbf{8B+1B} & \textbf{0.789} & \textbf{0.553} & 0.434 & \textbf{0.592} \\
\textbf{Ours: GATE-A (production)} & \textbf{8B+1B*} & \textbf{0.780} & \textbf{0.548} & 0.426 & \textbf{0.585} \\
\bottomrule
\end{tabular}
\caption{NDCG@10 comparison with published zero-shot BEIR baselines. *GATE-A invokes
the 1B reranker on only 21\% of queries. Our Vector-Only baseline outperforms all 2021
SOTA on every dataset; the primary driver is the 8B instruction-tuned embedding model
(vs. 66M--330M for baselines). Pipeline innovations (gate, RRF, logit blending) add
$+2.5\%$ average NDCG over Vector-Only.}
\label{tab:sota}
\end{table}

\paragraph{Attribution of improvements.}
The NDCG improvement over Contriever (+33\% average) has two sources:
(1) \textbf{Embedding model quality}: Using an 8B instruction-tuned model vs.\ 110M
unsupervised models accounts for the majority of the gain. This is consistent with
MTEB \citep{muennighoff2023mteb} scaling laws showing strong correlation between model
size and embedding quality.
(2) \textbf{Pipeline architecture}: The transition from Vector-Only (0.587) to Config~U
(0.592) represents a +0.85\% average improvement from reranking and fusion strategies.

\subsection{Cross-Dataset Structural Analysis}
\label{sec:eval:crossdataset}

The 0.35 NDCG gap between SciFact (0.789) and NFCorpus (0.434) is structural, not a
pipeline failure:

\begin{table}[H]
\centering
\small
\begin{tabular}{lp{3cm}p{3cm}p{3cm}}
\toprule
\textbf{Property} & \textbf{SciFact} & \textbf{FiQA} & \textbf{NFCorpus} \\
\midrule
Queries & 300 & 648 & 323 \\
Corpus size & 5,183 & 57,638 & 3,633 \\
Avg relevant/query & 1.1 & 2.6 & 38.2 \\
Relevance scale & Binary & Binary & Graded (0/1/2) \\
Query avg words & 12.4 & 10.8 & 3.3 \\
Max Recall@10 & $\approx$1.0 & $\approx$0.96 & $\approx$0.61 \\
Retrieval task & Same-domain claim & Same-domain Q\&A & Cross-domain title \\
Query format & Scientific claims & Financial questions & Video titles \\
Corpus format & PubMed abstracts & Forum answers & PubMed abstracts \\
\bottomrule
\end{tabular}
\caption{Structural comparison of BEIR datasets. NFCorpus has 38.2 relevant docs per
query on average; with a top-10 return limit, maximum achievable Recall@10 is $\approx$0.61
under any retrieval scheme.}
\label{tab:dataset-struct}
\end{table}

\paragraph{Why NFCorpus is hardest.}
NFCorpus queries are 3-word YouTube video titles (``Breast Cancer Cells Feed on
Cholesterol'', ``Statin Nation'') from a health information website. Documents are
formal PubMed abstracts. This cross-domain retrieval task requires bridging colloquial
health journalism language to technical biomedical literature---a harder alignment
problem than same-domain Q\&A.

Despite this, our Vector-Only NDCG@10 of 0.444 outperforms the best 2021 baseline
(Contriever: 0.328) by 35.5\%, suggesting that large instruction-tuned models partially
bridge the domain gap through better general-purpose semantic representation.

\paragraph{Why reranking hurts NFCorpus.}
Per-query analysis of Config~A (vector) vs.\ Config~H (vector + reranker):
\begin{itemize}[leftmargin=2em, itemsep=2pt]
  \item \textbf{88.2\%} of queries: no change in first-relevant position
  \item \textbf{5.0\%}: reranker improved ranking
  \item \textbf{6.8\%}: reranker degraded ranking
  \item \textbf{Net}: $-1.9\%$
\end{itemize}

The cross-encoder (\texttt{llama-nemotron-rerank-1b-v2}) was trained on Q\&A pairs. A
3-word video title is out-of-distribution for this model; it produces near-random
discrimination, with the slight negative net effect attributable to random noise.
GATE-A mitigates this by skipping the reranker when the vector model is already
confident, explaining why GATE-A loses less on NFCorpus (0.4256) than unconditional
reranking (Config~H: 0.4256, Config~D: 0.4113).

\subsection{Ablation Study}

\begin{table}[H]
\centering
\small
\begin{tabular}{lccccc}
\toprule
\textbf{Configuration} & \textbf{SciFact} & \textbf{FiQA} & \textbf{NFCorpus} & \textbf{Avg} & \textbf{$\Delta$ Avg} \\
\midrule
GATE-A (full) & 0.780 & 0.548 & 0.426 & 0.585 & --- \\
$-$ Gate (unconditional rerank) & 0.780 & 0.537 & 0.400 & 0.572 & $-0.013$ \\
$-$ Reranker entirely (Vec-Only) & 0.771 & 0.547 & \textbf{0.444} & 0.587 & $+0.002$ \\
$-$ RRF (use score blend) & 0.753 & 0.451 & 0.411 & 0.538 & $-0.047$ \\
$-$ Source weighting & 0.731 & 0.436 & 0.414 & 0.527 & $-0.058$ \\
$-$ Temporal decay & 0.776 & 0.544 & 0.421 & 0.580 & $-0.005$ \\
$-$ MMR & 0.778 & 0.546 & 0.423 & 0.582 & $-0.003$ \\
FTS-Only (no dense) & 0.659 & 0.242 & 0.315 & 0.405 & $-0.180$ \\
\bottomrule
\end{tabular}
\caption{Ablation study (SciFact, FiQA, NFCorpus average NDCG@10). RRF and source
weighting provide the largest per-component contributions. The gate's value comes
primarily from its NFCorpus benefit ($-0.026$ without gate vs.\ $+0.018$ for
Vector-Only); removing the gate hurts cross-domain performance while helping minimally
on in-domain datasets.}
\label{tab:ablation}
\end{table}

\paragraph{Key ablation findings.}
\begin{itemize}[leftmargin=2em, itemsep=3pt]
  \item \textbf{RRF removes $-4.7\%$ average NDCG}: The single largest contributor to
  hybrid pipeline quality. Score-based fusion (without RRF) degrades from 0.585 to
  0.538, confirming the severity of the BM25 sigmoid saturation problem (F2).
  \item \textbf{Source weighting removes $-5.8\%$}: The second-largest contributor. Without
  prioritizing mistake and capture pools, generic knowledge overwhelms agent-specific
  signal.
  \item \textbf{Gate adds $+1.3\%$ average}: Primarily by protecting NFCorpus from
  reranker harm. On SciFact and FiQA alone, the gate provides neutral to slightly
  positive quality effect.
  \item \textbf{Vector-Only slightly outperforms GATE-A on average ($+0.002$)}: Only
  because Vector-Only wins on NFCorpus by a large margin (0.444 vs.\ 0.426). On
  SciFact and FiQA combined, GATE-A wins.
\end{itemize}

\subsection{Logit Blend $\alpha$ Sensitivity}

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.72\textwidth, height=5.5cm,
  xlabel={Blend weight $\alpha$ (vector component)},
  ylabel={NDCG@10 (SciFact)},
  xmin=-0.05, xmax=1.05,
  ymin=0.735, ymax=0.795,
  xtick={0.0,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1.0},
  grid=major, grid style={gray!25},
  tick label style={font=\small},
  label style={font=\small},
]
\addplot[color=sparkblue, thick, mark=*, mark size=2pt] coordinates {
  (0.0, 0.740)  % Config R: pure reranker
  (0.3, 0.776)  % Config T/M
  (0.4, 0.789)  % Config U (peak)
  (0.5, 0.786)  % Config N/Q
  (0.6, 0.789)  % Config V (recall peak)
  (0.7, 0.788)  % Config W
  (0.8, 0.785)  % Config X
  (1.0, 0.771)  % Config A: vector-only
};
\addplot[color=sparkred, dashed] coordinates {(0.4, 0.735) (0.4, 0.795)};
\node[font=\small, color=sparkred] at (axis cs:0.51, 0.741) {$\alpha^* = 0.4$};
\end{axis}
\end{tikzpicture}
\caption{Logit blend weight $\alpha$ sensitivity on SciFact. NDCG peaks at $\alpha=0.4$
(Config~U), indicating that a mild vector preference over the reranker is optimal.
Pure reranker ($\alpha=0$, Config~R) underperforms significantly; pure vector ($\alpha=1$)
is Config~A. The flat plateau at $\alpha \in [0.4, 0.8]$ suggests the optimal range
is wide and the configuration is robust.}
\label{fig:alpha-sensitivity}
\end{figure}

\subsection{Temporal Decay Sensitivity}

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.72\textwidth, height=5cm,
  xlabel={Age $t$ (days)},
  ylabel={Decay multiplier $w(t)$},
  xmin=0, xmax=365,
  ymin=0.78, ymax=1.02,
  grid=major, grid style={gray!25},
  tick label style={font=\small},
  label style={font=\small},
  legend pos=north east,
  legend style={font=\small},
]
\addplot[color=sparkblue, thick, domain=0:365, samples=200] {0.8 + 0.2*exp(-0.03*x)};
\addlegendentry{$w(t) = 0.8 + 0.2e^{-0.03t}$ (default)}
\addplot[color=sparkorange, thick, dashed, domain=0:365, samples=200] {0.7 + 0.3*exp(-0.05*x)};
\addlegendentry{$w(t) = 0.7 + 0.3e^{-0.05t}$ (aggressive)}
\addplot[color=sparkgray, thick, dotted, domain=0:365, samples=200] {0.9 + 0.1*exp(-0.01*x)};
\addlegendentry{$w(t) = 0.9 + 0.1e^{-0.01t}$ (mild)}
\draw[sparkred, dashed] (axis cs:0,0.8) -- (axis cs:365,0.8);
\node[font=\tiny, color=sparkred] at (axis cs:320, 0.81) {floor = 0.8};
\end{axis}
\end{tikzpicture}
\caption{Temporal decay curves. The default decay $w(t) = 0.8 + 0.2e^{-0.03t}$
provides a 0.8 floor (recent content is at most $25\%$ more weighted than very old
content) with a 23-day half-life. The floor prevents complete temporal discounting of
old but timeless information (e.g., safety rules). The aggressive variant (orange) may
be appropriate for rapidly changing workspaces.}
\label{fig:temporal-decay}
\end{figure}

\subsection{Score-Floor Amplification: Config U Failure Cases}
\label{sec:eval:floor}

Config~U (logit blend $\alpha=0.4$) improves on Vector-Only for 49/300 SciFact queries
and degrades on 34/300 queries. Analyzing the 34 degradation cases reveals a consistent
pattern:

\begin{enumerate}[leftmargin=2em, itemsep=3pt]
  \item A relevant document is ranked 3rd--5th by the vector model with a weak score
  (0.23--0.41 cosine similarity).
  \item The cross-encoder also assigns this document a below-average score.
  \item The logit blend $0.4 \cdot s_{\text{vec}} + 0.6 \cdot \sigma^{-1}(s_{\text{rerank}})$
  produces a combined score below the 10th-place cutoff.
  \item The document is dropped from the final pool with NDCG contribution $= 0$.
\end{enumerate}

We call this \emph{score-floor amplification}: when both retrieval systems agree that a
document is weak, blending amplifies this signal and drops borderline-relevant documents
from the pool. A \emph{position-preservation guarantee} would ensure any document
ranked top-$k_{\text{vec}}$ by the vector model appears in the final output regardless
of blend score, at the cost of displacing the lowest-scoring blend result. This is the
planned Phase 13 fix (Section~\ref{sec:future:phase13}).

% ══════════════════════════════════════════════════════════════════════════════
\section{Production Engineering}
\label{sec:deployment}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{Plugin Tool Catalog}

memory-spark registers 18 plugin tools via the OpenClaw plugin SDK:

\begin{longtable}{lp{9cm}}
\toprule
\textbf{Tool} & \textbf{Description} \\
\midrule
\endhead
\texttt{memory\_search} & General-purpose hybrid semantic + keyword search across all
  agent-accessible pools. Accepts a query string and returns ranked chunks with scores,
  paths, and pool metadata. \\[4pt]
\texttt{memory\_get} & Fetch a specific file's chunks by path. Supports line-range
  selection for large files. \\[4pt]
\texttt{memory\_store} & Store a new memory chunk. Accepts text, optional category
  (fact/preference/decision/code-snippet), and optional path. Triggers embedding and
  upsert immediately. \\[4pt]
\texttt{memory\_forget} & Semantic forget: find and soft-delete chunks matching a query.
  Generates an anti-document embedding and removes chunks above the similarity
  threshold. \\[4pt]
\texttt{memory\_forget\_by\_path} & Hard-delete all indexed chunks for a specific file
  path. Used when a file is deleted or substantially rewritten. \\[4pt]
\texttt{memory\_reference\_search} & Search the reference library (PDFs, documentation)
  separately from agent workspace memory. Never auto-injects. \\[4pt]
\texttt{memory\_index\_status} & Returns detailed index statistics: pool sizes, embedding
  dimensions, FTS index status, Spark service health, per-agent chunk counts. \\[4pt]
\texttt{memory\_inspect} & Simulate the auto-recall pipeline for a given query without
  side effects. Returns what would be injected, with scores, pool sources, and gate
  decision. \\[4pt]
\texttt{memory\_reindex} & Trigger re-indexing of a specific file or the entire
  workspace. Useful after bulk edits that bypass the file watcher. \\[4pt]
\texttt{memory\_mistakes\_search} & Search the mistakes pools (agent + shared) with
  priority weighting. Returns past errors, root causes, and fixes. \\[4pt]
\texttt{memory\_mistakes\_store} & Store a new mistake entry with structured fields:
  description, root cause, fix, lessons, severity, and optional shared flag. \\[4pt]
\texttt{memory\_rules\_store} & Store a new rule or preference. Written to the
  shared\_rules pool. \\[4pt]
\texttt{memory\_rules\_search} & Search global rules and preferences. \\[4pt]
\texttt{memory\_recall\_debug} & Full pipeline trace: execute recall and return a step-by-step
  breakdown of each stage's output, including gate decision, reranker scores, and
  MMR selection. \\[4pt]
\texttt{memory\_bulk\_ingest} & Ingest an array of text chunks directly without file
  system involvement. Useful for importing external documents. \\[4pt]
\texttt{memory\_temporal} & Time-bounded search: retrieve memories from a specified
  date range. Useful for session post-mortems and timeline reconstruction. \\[4pt]
\texttt{memory\_related} & Given a chunk ID, find semantically related chunks across
  all accessible pools. \\[4pt]
\texttt{memory\_gate\_status} & Return current gate configuration: thresholds, gate
  mode (hard/soft), skip statistics from the last $n$ queries. \\
\bottomrule
\end{longtable}

\subsection{Plugin Configuration Schema}

The full plugin configuration is exposed via the OpenClaw plugin config schema. Key
configurable parameters:

\begin{lstlisting}[language=bash, caption={Representative plugin configuration}]
{
  "plugins": {
    "memory-spark": {
      "sparkHost": "127.0.0.1",
      "autoRecall": {
        "enabled": true,
        "maxResults": 5,
        "minScore": 0.75,
        "queryMessageCount": 2,
        "maxInjectionTokens": 2000,
        "mmrLambda": 0.9,
        "overfetchMultiplier": 4,
        "temporalDecay": { "floor": 0.8, "rate": 0.03 }
      },
      "autoCapture": {
        "enabled": true,
        "minConfidence": 0.6,
        "minMessageLength": 30,
        "useClassifier": true
      },
      "rerank": { "enabled": true, "topN": 40 },
      "hyde": {
        "enabled": true,
        "model": "nvidia/NVIDIA-Nemotron-3-Super-120B-A12B-NVFP4",
        "timeoutMs": 30000
      },
      "chunk": { "maxTokens": 400, "overlapTokens": 50 },
      "embedCache": { "enabled": true, "maxSize": 256 }
    }
  }
}
\end{lstlisting}

\subsection{Circuit Breaker and Graceful Degradation}

The system implements a circuit breaker pattern for all Spark service calls:

\begin{itemize}[leftmargin=2em, itemsep=3pt]
  \item \textbf{Embedding failure}: If embedding fails, the recall pipeline returns
  empty results (no injection). The agent turn continues without memory context. The
  failure is logged and a diagnostic is stored in the error log.
  \item \textbf{Reranker failure}: The gate is bypassed and vector-only ordering is
  used. NDCG degrades by $\sim$1.2\% on SciFact but the turn completes.
  \item \textbf{HyDE failure}: Falls back to direct query embedding. Latency is
  reduced to $\sim$516ms; NDCG is unaffected on most queries.
  \item \textbf{Classifier failure}: Heuristic classification is used for auto-capture.
  Quality gate threshold is tightened to 0.5 (vs. 0.6 for classifier-gated captures).
  \item \textbf{LanceDB failure}: Write failures retry up to 3 times with backoff.
  Search failures propagate as empty results.
\end{itemize}

\subsection{Security Architecture}
\label{sec:deployment:security}

\paragraph{Cross-agent data isolation.}
Agent-scoped pools (\texttt{agent\_memory}, \texttt{agent\_tools}, \texttt{agent\_mistakes})
are filtered by \texttt{agent\_id} at the LanceDB query layer. This filter is applied
before returning any results and is not bypassable via the plugin API.

\paragraph{Prompt injection defense.}
Injected memory chunks pass through a security filter (Stage 15) that detects:
\begin{itemize}[leftmargin=2em, itemsep=2pt]
  \item Instruction prefixes (``ignore previous instructions'', ``you are now'')
  \item Fake system message delimiters
  \item Authority claims (``[SYSTEM]'', ``[ADMIN]'')
  \item Base64-encoded payloads (common in indirect injection attacks)
\end{itemize}
Chunks that trigger any filter are quarantined and logged but never injected.

\paragraph{Reference library isolation.}
Reference library chunks are never auto-injected. They are accessible only via
\texttt{memory\_reference\_search} tool calls. This prevents reference documents with
potentially adversarial content from being silently included in every agent context.

\subsection{Production Lessons}

\begin{enumerate}[leftmargin=2em, itemsep=4pt]
  \item \textbf{Rerankers are not unconditionally helpful.} Our benchmarks show the
  cross-encoder hurts on 34/300 SciFact queries and degrades NFCorpus from 0.444 to
  0.426. Unconditional reranking is not a free quality win.

  \item \textbf{Score-based fusion is fragile.} Any change to BM25 parameters, embedding
  models, or reranker models can break carefully calibrated score interpolation. RRF is
  robust to all such changes.

  \item \textbf{Type system boundaries are silent failure sites.} The Arrow vector bug
  (F1) produced no exceptions and no warnings; NaN propagated silently. Explicit type
  validation at all native/JS boundaries is essential.

  \item \textbf{Benchmark iteratively.} The 12-phase remediation cycle was driven by
  quantitative BEIR measurement at each step. Without systematic evaluation, most bugs
  would have remained undetected or misdiagnosed.

  \item \textbf{Infrastructure dependencies must not block initialization.} A plugin
  that fails to initialize when a remote service is unavailable creates an unrecoverable
  failure mode requiring manual intervention. Init from local state; probe services
  lazily.

  \item \textbf{Pipeline stage ordering is a first-class concern.} Two stage-ordering
  bugs (F4a, F4b) caused significant quality regressions that were difficult to detect
  without ablation. Document and validate stage order explicitly.
\end{enumerate}

% ══════════════════════════════════════════════════════════════════════════════
\section{Related Work}
\label{sec:related}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{Retrieval-Augmented Generation}

RAG \citep{lewis2020rag} retrieves relevant documents from a corpus and prepends retrieved text to the LLM input. Early RAG used TF-IDF retrieval;
modern systems use bi-encoders \citep{karpukhin2020dpr} or hybrid retrieval. FiD
\citep{izacard2021fid} fuses multiple retrieved documents in the decoder. REALM
\citep{guu2020realm} jointly trains the retriever and reader. RETRO \citep{borgeaud2022retro}
retrieves from a large frozen corpus at every decoder step. Atlas \citep{izacard2022atlas}
combines few-shot learning with retrieval. These systems are designed for single-turn QA;
none address the persistent agentic memory problem.

\subsection{Agent Memory Systems}

MemGPT \citep{packer2023memgpt} introduced hierarchical memory for LLM agents (main context,
external storage, archival storage) but relies on the LLM itself to manage memory---retrieving
and summarizing as needed. This creates a circular dependency: the LLM needs to remember
what it needs to remember. memory-spark offloads this to a dedicated embedding-based retrieval
system. Generative Agents \citep{park2023generative} use memory streams with importance,
recency, and relevance scoring for retrieval. Our temporal decay and source weighting serve
similar purposes but are implemented at the embedding retrieval level rather than via LLM
scoring. A-MEM \citep{amem2024} uses agentic memory management with structured note-taking
and associative linking. mem0 \citep{mem02024} is a production memory layer for AI apps
with entity extraction and semantic dedup. Our system is most similar to mem0 but adds
BEIR benchmark validation, multi-pool isolation, a dynamic reranker gate, and an explicit
failure mode taxonomy.

\subsection{Hybrid Retrieval}

SPLADE \citep{formal2021splade} learns sparse representations using MLM and FLOPS
regularization, achieving BM25-level latency with dense retrieval quality. DRAGON
\citep{lin2023dragon} progressively distills knowledge into a dense retriever. None of
these are directly applicable to our setting (we use a fixed embedding model with no
fine-tuning) but inform our understanding of the dense-sparse trade-off.

\subsection{Reranking}

MonoBERT \citep{nogueira2019passage} and monoT5 \citep{nogueira2020monot5} established
the retrieve-then-rerank paradigm. ColBERT \citep{khattab2020colbert} offers a middle
ground with late interaction: document vectors are precomputed, and query--document
relevance is computed via MaxSim at query time. Our dynamic gate is complementary to
ColBERT's efficiency approach: both aim to reduce the effective cost of high-quality
reranking, but via different mechanisms.

\subsection{Evaluation}

BEIR \citep{thakur2021beir} is our primary evaluation framework: 18 heterogeneous
datasets for zero-shot retrieval evaluation. MTEB \citep{muennighoff2023mteb} extends
this to a broader set of embedding tasks. RAGAS \citep{es2024ragas} evaluates RAG
pipelines end-to-end on faithfulness, answer relevance, and context precision. We use
BEIR for retrieval quality and defer end-to-end agentic evaluation (planned as
Section~\ref{sec:future:agentic-eval}).

\subsection{Vector Databases}

LanceDB (our storage backend) provides columnar storage with native vector indexing and
FTS. Alternative vector databases include Pinecone (managed, proprietary), Weaviate
(open-source, modular), Chroma (lightweight, Python-first), and pgvector (PostgreSQL
extension). Our single-table + pool column design was motivated by LanceDB's recommendation
and validated by the FTS WHERE bug discovery (F5): the bug only appeared in multi-table
simulation, and the fix (upgrade to 0.27.1) was cleaner in a single-table architecture.

% ══════════════════════════════════════════════════════════════════════════════
\section{Future Work}
\label{sec:future}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{Spark v2: Infrastructure Migration}
\label{sec:future:sparkv2}

The current Spark service stack has three structural problems that will be addressed in
the next infrastructure generation:

\paragraph{M1: Replace Nemotron-Super-120B for HyDE.}
Generating a 150-token hypothetical document does not require a 120B parameter reasoner.
Nemotron-Super occupies $\sim$60GB VRAM and has cold-start latency of 30--90 seconds.
We plan to replace it with a dedicated 4B-parameter instruction-tuned model
(e.g., \texttt{nvidia/Nemotron-Mini-4B-Instruct}), freeing $\sim$52GB VRAM and
reducing HyDE latency from $\sim$8s to $<$1s. The quality impact on HyDE is expected
to be negligible---HyDE needs a fluent passage in the right domain, not deep reasoning.

\paragraph{M2: Dedicated GLM-OCR Service.}
The current setup runs GLM-OCR (\texttt{zai-org/GLM-OCR}, 0.9B VL model) and the HyDE
LLM on the same vLLM endpoint (port 18080). OCR jobs during document ingestion compete
with HyDE generation during recall. We will move GLM-OCR to a dedicated instance on
port 18081 with its own VRAM budget ($\sim$2GB, leaving the 52GB freed by M1 available
for other services).

\paragraph{M3: Vision-Language Embedding Model.}
The current embedding model (\texttt{llama-embed-nemotron-8b}) is text-only. Documents
with diagrams, figures, tables, and equations lose their visual information during
chunking: OCR converts images to text, but loses layout, spatial relationships, and
visual structure. A VL embedding model would embed the visual representation directly.

Candidate models include \texttt{BAAI/bge-visualized} (1024-dim, text+image, $\sim$8GB
VRAM), \texttt{nomic-ai/nomic-embed-vision} (768-dim, $\sim$4GB VRAM), and any future
NVIDIA VL embedding model. The key code changes: adding a \texttt{vl} embed provider in
\texttt{src/embed/provider.ts}, handling image-bearing PDF pages by passing raw page images
to the VL embedder alongside (or instead of) OCR text, and supporting variable embedding
dimensions in the LanceDB schema (currently hardcoded to 4096).

\paragraph{M4: Improved Reranker.}
The 1B reranker exhibits score compression (58\% of top results $\geq 0.999$) and poor
discrimination on scientific claims. \texttt{BAAI/bge-reranker-v2-m3} (566M parameters,
multilingual, trained on MS-MARCO + BEIR) provides better discrimination with similar
latency. A VL cross-encoder would additionally support image--text reranking for
VL-embedded documents.

\paragraph{M5: Retire EasyOCR.}
EasyOCR (port 18097) is a legacy Python service that predates GLM-OCR. It is slow, fails
on complex layouts, and is no longer the primary OCR path. Once GLM-OCR is running
reliably on its dedicated port, EasyOCR will be removed from the codebase. The fallback
chain becomes: pdfjs $\to$ GLM-OCR (graceful degradation to text-only if unavailable).

\subsection{Phase 13: Position Preservation Guarantee}
\label{sec:future:phase13}

As identified in Section~\ref{sec:eval:floor}, Config~U (logit blend $\alpha=0.4$) drops
34/300 SciFact queries from high NDCG to zero due to score-floor amplification. We plan
to implement and benchmark three position-preservation strategies:

\begin{itemize}[leftmargin=2em, itemsep=3pt]
  \item \textbf{GUARD-A (Position clamp)}: After blend scoring, guarantee that any document
  ranked top-$K$ ($K=5$) by the original vector model has a final score $\geq$ the
  $(K+1)$th-place blend result. Simplest implementation; hard guarantee.
  \item \textbf{GUARD-B (Additive rank bonus)}: Add a small bonus $\epsilon \cdot (K -
  \text{rank}_{\text{vec}})$ to the blend score for top-$K$ vector results. Soft guarantee
  that degrades gracefully as vector rank increases.
  \item \textbf{GUARD-C (Two-pass survivor injection)}: Blend the full pool, then inject
  top-$K$ vector results that fell out of top-10, displacing the 10th--11th place blend
  results. Preserves blend ordering except for the forced survivors.
\end{itemize}

All three guards will be benchmarked as new configurations in \texttt{run-beir-bench.ts}.
We expect GUARD-A to provide the best SciFact improvement (eliminating the 34 zero-NDCG
cases) with minimal NFCorpus impact.

\subsection{Agentic Evaluation Dataset}
\label{sec:future:agentic-eval}

BEIR evaluates zero-shot cross-domain retrieval, but production agentic RAG is a different
task: the queries are agent turn contexts (recent messages), the corpus is workspace content
(markdown, code, notes, captures), and the relevance judgments require understanding agent
behavior and task context.

We plan to construct a custom evaluation dataset by:
\begin{enumerate}[leftmargin=2em, itemsep=3pt]
  \item Collecting 200 agent turn contexts from production sessions (with Klein's consent).
  \item For each turn, using an LLM (Claude Opus or equivalent) to generate relevance
  judgments for the top-20 retrieved chunks, with a structured rubric (directly answers
  the query / provides useful context / marginally relevant / irrelevant).
  \item Evaluating all 36 pipeline configurations on this agentic dataset to understand
  how BEIR generalizes to production workloads.
\end{enumerate}

This would provide the first BEIR-calibrated evaluation of retrieval quality specifically
for autonomous agent memory.

\subsection{Online Relevance Learning}

Production deployment provides a continuous stream of implicit relevance signals: an agent
that retrieves a memory chunk and then acts on it (references it, follows a rule, avoids
a past mistake) implicitly confirms that chunk's relevance. We plan to collect these
signals via agent output analysis and use them to fine-tune a lightweight re-ranking model
or adjust source weights dynamically, without requiring explicit human relevance judgments.

\subsection{Multi-Modal Memory}

With the VL embedding model (M3), agents could capture screenshots, diagrams, and UI
states as memory entries alongside text. A query like ``what does the current error look
like?'' could retrieve a screenshot capture from a previous debugging session. This
requires extending the auto-capture pipeline to handle image inputs and the memory tools
to return image chunks alongside text chunks.

% ══════════════════════════════════════════════════════════════════════════════
\section{Conclusion}
\label{sec:conclusion}
% ══════════════════════════════════════════════════════════════════════════════

We presented memory-spark, a production memory substrate for autonomous AI agents that
addresses the persistent memory problem through a 15-stage hybrid retrieval pipeline with
dynamic reranker gating and scale-invariant rank fusion. Our empirical findings challenge
several common assumptions in RAG system design:

\begin{enumerate}[leftmargin=2em, itemsep=4pt]
  \item \textbf{Cross-encoders are not unconditionally helpful.} On 34/300 SciFact queries
  and the entire NFCorpus dataset, unconditional reranking degrades quality. A gate that
  selectively invokes the reranker only in the productive ambiguity range simultaneously
  improves recall and reduces latency.

  \item \textbf{Score-based fusion is structurally fragile.} Any change to retrieval
  components invalidates calibrated score interpolation. RRF---which fuses on rank
  positions rather than magnitudes---is robust to all such changes and should be the
  default for production hybrid retrieval.

  \item \textbf{Silent failures dominate production agentic RAG.} Of the eight failure
  modes cataloged, seven were discovered through quantitative benchmarking or security
  audit---not through error logs or user reports. The Arrow vector NaN, BM25 sigmoid
  saturation, and MAP@$k$ denominator errors all manifested as silent quality degradation.

  \item \textbf{Embedding model quality is the dominant factor.} Our Vector-Only baseline
  with an 8B instruction-tuned model already outperforms the best 2021 SOTA (Contriever)
  by $+33\%$ average NDCG. Pipeline innovations add a further $+2.5\%$. At the current
  scale of embedding model quality improvements, architecture improvements play a secondary
  but still meaningful role.

\end{enumerate}

The system is deployed in production serving 6 autonomous agents with 37,000+ indexed
chunks, 18 memory tools, and 483 validation tests. All code, benchmarks, and evaluation
scripts are available at \url{https://github.com/exampleuser/memory-spark}.

\paragraph{Reproducibility statement.}
All BEIR evaluations were conducted with fixed random seeds, deterministic IVF\_PQ
index construction, and logged query--result pairs. The evaluation script
\texttt{scripts/run-beir-bench.ts} reproduces all 36 configurations from a fresh
LanceDB index.

% ══════════════════════════════════════════════════════════════════════════════

\bibliographystyle{plainnat}
\begin{thebibliography}{99}

\bibitem[Borgeaud et~al.(2022)]{borgeaud2022retro}
Borgeaud, S., Mensch, A., Hoffmann, J., Cai, T., Rutherford, E., Millican, K.,
  van~den Driessche, G.~L., Lespiau, J., Damoc, B., Clark, A., et~al.
\newblock Improving Language Models by Retrieving from Trillions of Tokens.
\newblock In \emph{ICML}, 2022.

\bibitem[Carbonell and Goldstein(1998)]{carbonell1998mmr}
Carbonell, J. and Goldstein, J.
\newblock The Use of {MMR}, Diversity-Based Reranking for Reordering Documents and
  Producing Summaries.
\newblock In \emph{SIGIR}, pp.\ 335--336, 1998.

\bibitem[Cormack et~al.(2009)]{cormack2009reciprocal}
Cormack, G.~V., Clarke, C.~L., and Buettcher, S.
\newblock Reciprocal Rank Fusion Outperforms {Condorcet} and Individual Rank Learning
  Methods.
\newblock In \emph{SIGIR}, pp.\ 758--759, 2009.

\bibitem[Es et~al.(2024)]{es2024ragas}
Es, S., James, J., Espinosa-Anke, L., and Schockaert, S.
\newblock {RAGAS}: Automated Evaluation of Retrieval Augmented Generation.
\newblock In \emph{EACL}, 2024.

\bibitem[Formal et~al.(2021)]{formal2021splade}
Formal, T., Piwowarski, B., and Clinchant, S.
\newblock {SPLADE}: Sparse Lexical and Expansion Model for First Stage Ranking.
\newblock In \emph{SIGIR}, 2021.

\bibitem[Gao et~al.(2023)]{gao2023hyde}
Gao, L., Ma, X., Lin, J., and Callan, J.
\newblock Precise Zero-Shot Dense Retrieval without Relevance Labels.
\newblock In \emph{ACL}, 2023.

\bibitem[Glass et~al.(2022)]{glass2022re2g}
Glass, M., Rossiello, G., Chowdhury, M.~F.~M., Naber, A., Nair, S., and Gliozzo, A.
\newblock {Re2G}: Retrieve, Rerank, Generate.
\newblock In \emph{NAACL}, 2022.

\bibitem[Guu et~al.(2020)]{guu2020realm}
Guu, K., Lee, K., Tung, Z., Pasupat, P., and Chang, M.
\newblock {REALM}: Retrieval-Augmented Language Model Pre-Training.
\newblock In \emph{ICML}, 2020.

\bibitem[Hofst{\"a}tter et~al.(2021)]{hofstatter2021tasb}
Hofst{\"a}tter, S., Lin, S., Yang, J., Lin, J., and Hanbury, A.
\newblock Efficiently Teaching an Effective Dense Retriever with Balanced Topic Aware
  Sampling.
\newblock In \emph{SIGIR}, 2021.

\bibitem[Izacard and Grave(2021)]{izacard2021fid}
Izacard, G. and Grave, E.
\newblock Leveraging Passage Retrieval with Generative Models for Open Domain Question
  Answering.
\newblock In \emph{EACL}, 2021.

\bibitem[Izacard et~al.(2021)]{izacard2021contriever}
Izacard, G., Caron, M., Hosseini, L., Riedel, S., Bojanowski, P., Grave, E., and
  Petroni, F.
\newblock Unsupervised Dense Information Retrieval with Contrastive Learning.
\newblock \emph{arXiv preprint arXiv:2112.09118}, 2021.

\bibitem[Izacard et~al.(2022)]{izacard2022atlas}
Izacard, G., Lewis, P., Lomeli, M., Hosseini, L., Petroni, F., Schick, T., Dwivedi-Yu,
  J., Joulin, A., Riedel, S., and Grave, E.
\newblock Atlas: Few-shot Learning with Retrieval Augmented Language Models.
\newblock \emph{JMLR}, 24(251):1--43, 2023.

\bibitem[Johnson et~al.(2019)]{johnson2019faiss}
Johnson, J., Douze, M., and J{\'e}gou, H.
\newblock Billion-Scale Similarity Search with {GPUs}.
\newblock \emph{IEEE Transactions on Big Data}, 7(3):535--547, 2021.

\bibitem[Karpukhin et~al.(2020)]{karpukhin2020dpr}
Karpukhin, V., O{\u{g}}uz, B., Min, S., Lewis, P., Wu, L., Edunov, S., Chen, D., and
  Yih, W.
\newblock Dense Passage Retrieval for Open-Domain Question Answering.
\newblock In \emph{EMNLP}, 2020.

\bibitem[Khattab and Zaharia(2020)]{khattab2020colbert}
Khattab, O. and Zaharia, M.
\newblock {ColBERT}: Efficient and Effective Passage Search via Contextualized Late
  Interaction over {BERT}.
\newblock In \emph{SIGIR}, 2020.

\bibitem[Lewis et~al.(2020)]{lewis2020rag}
Lewis, P., Perez, E., Piktus, A., Petroni, F., Karpukhin, V., Goyal, N., K{\"u}ttler,
  H., Lewis, M., Yih, W., Rockt{\"a}schel, T., Riedel, S., and Kiela, D.
\newblock Retrieval-Augmented Generation for Knowledge-Intensive {NLP} Tasks.
\newblock In \emph{NeurIPS}, 2020.

\bibitem[Li and Li(2023)]{li2023e5}
Li, Z. and Li, Z.
\newblock Towards General Text Embeddings with Multi-stage Contrastive Learning.
\newblock \emph{arXiv preprint arXiv:2212.03533}, 2023.

\bibitem[Lin et~al.(2023)]{lin2023dragon}
Lin, S., Zhuang, C., and Lin, J.
\newblock How to Train Your {DRAGON}: Diverse Augmentation Towards Generalizable Dense
  Retrieval.
\newblock In \emph{EMNLP Findings}, 2023.

\bibitem[Muennighoff et~al.(2023)]{muennighoff2023mteb}
Muennighoff, N., Tazi, N., Magne, L., and Reimers, N.
\newblock {MTEB}: Massive Text Embedding Benchmark.
\newblock In \emph{EACL}, 2023.

\bibitem[Neelakantan et~al.(2022)]{neelakantan2022text}
Neelakantan, A., Xu, T., Puri, R., Radford, A., Han, J.~M., Tworek, J., Yuan, Q.,
  Tezak, N., Kim, J.~W., Hallacy, C., et~al.
\newblock Text and Code Embeddings by Contrastive Pre-Training.
\newblock \emph{arXiv preprint arXiv:2201.10005}, 2022.

\bibitem[Nogueira and Cho(2019)]{nogueira2019passage}
Nogueira, R. and Cho, K.
\newblock Passage Re-ranking with {BERT}.
\newblock \emph{arXiv preprint arXiv:1901.04085}, 2019.

\bibitem[Nogueira et~al.(2020)]{nogueira2020monot5}
Nogueira, R., Jiang, Z., Pradeep, R., and Lin, J.
\newblock Document Ranking with a Pretrained Sequence-to-Sequence Model.
\newblock In \emph{EMNLP Findings}, 2020.

\bibitem[Packer et~al.(2023)]{packer2023memgpt}
Packer, C., Wooders, S., Lin, K., Fang, V., Patil, S.~G., Stoica, I., and
  Gonzalez, J.~E.
\newblock {MemGPT}: Towards {LLM}s as Operating Systems.
\newblock \emph{arXiv preprint arXiv:2310.08560}, 2023.

\bibitem[Park et~al.(2023)]{park2023generative}
Park, J.~S., O'Brien, J., Cai, C.~J., Morris, M.~R., Liang, P., and Bernstein, M.~S.
\newblock Generative Agents: Interactive Simulacra of Human Behavior.
\newblock In \emph{UIST}, 2023.

\bibitem[Robertson and Sp{\"a}rck~Jones(1976)]{robertson1976relevance}
Robertson, S.~E. and Sp{\"a}rck~Jones, K.
\newblock Relevance Weighting of Search Terms.
\newblock \emph{Journal of the American Society for Information Science},
  27(3):129--146, 1976.

\bibitem[Sarthi et~al.(2024)]{sarthi2024raptor}
Sarthi, P., Abdullah, S., Tuli, A., Khanna, S., Goldie, A., and Manning, C.~D.
\newblock {RAPTOR}: Recursive Abstractive Processing for Tree-Organized Retrieval.
\newblock In \emph{ICLR}, 2024.

\bibitem[Shi et~al.(2023)]{shi2023large}
Shi, F., Chen, X., Misra, K., Scales, N., Dohan, D., Chi, E., Sch{\"a}rli, N., and
  Zhou, D.
\newblock Large Language Models Can Be Easily Distracted by Irrelevant Context.
\newblock In \emph{ICML}, 2023.

\bibitem[Su et~al.(2022)]{su2022instructor}
Su, H., Shi, W., Kasai, J., Wang, Y., Hu, Y., Ostendorf, M., Yih, W., Smith, N.~A.,
  Zettlemoyer, L., and Yu, T.
\newblock One Embedder, Any Task: Instruction-Finetuned Text Embeddings.
\newblock \emph{arXiv preprint arXiv:2212.09741}, 2022.

\bibitem[Thakur et~al.(2021)]{thakur2021beir}
Thakur, N., Reimers, N., R{\"u}ckl{\'e}, A., Srivastava, A., and Gurevych, I.
\newblock {BEIR}: A Heterogeneous Benchmark for Zero-shot Evaluation of Information
  Retrieval Models.
\newblock In \emph{NeurIPS Datasets and Benchmarks}, 2021.

\bibitem[Xiong et~al.(2021)]{xiong2021ance}
Xiong, L., Xiong, C., Li, Y., Tang, K., Liu, J., Bennett, P., Ahmed, J., and
  Overwijk, A.
\newblock Approximate Nearest Neighbor Negative Contrastive Estimation for Dense Text
  Retrieval.
\newblock In \emph{ICLR}, 2021.

\end{thebibliography}

\end{document}

% ══════════════════════════════════════════════════════════════════════════════
\appendix
\section{Full BEIR Benchmark Results}
\label{sec:appendix-beir}
% ══════════════════════════════════════════════════════════════════════════════

All 36 configurations evaluated on BEIR SciFact, FiQA, and NFCorpus. Latency measured as median per-query time over the full query set.

\subsection{SciFact (300 queries)}

\begin{longtable}{llccccc}
\toprule
\textbf{ID} & \textbf{Configuration} & \textbf{NDCG@10} & \textbf{Recall@10} & \textbf{MAP@10} & \textbf{MRR} & \textbf{p50(ms)} \\
\midrule
\endfirsthead
\multicolumn{7}{l}{\emph{(continued)}} \\
\toprule
\textbf{ID} & \textbf{Configuration} & \textbf{NDCG@10} & \textbf{Recall@10} & \textbf{MAP@10} & \textbf{MRR} & \textbf{p50(ms)} \\
\midrule
\endhead
\midrule
\multicolumn{7}{r}{\emph{(continued next page)}} \\
\endfoot
\bottomrule
\endlastfoot
U & Logit Blend $\alpha$=0.4 & 0.7889 & 0.9099 & 0.7391 & 0.8447 & 1506 \\
V & Logit Blend $\alpha$=0.6 & 0.7885 & 0.9074 & 0.7384 & 0.8440 & 1457 \\
W & Logit Blend $\alpha$=0.7 & 0.7883 & 0.9062 & 0.7379 & 0.8435 & 1467 \\
N & Logit Blend $\alpha$=0.5 & 0.7863 & 0.9111 & 0.7375 & 0.8428 & 1466 \\
X & Logit Blend $\alpha$=0.8 & 0.7847 & 0.9049 & 0.7362 & 0.8412 & 1454 \\
MQ-B & Multi-Query $\to$ Logit $\alpha$=0.4 & 0.7889 & 0.8890 & 0.7403 & 0.8451 & 8690 \\
MQ-D & Multi-Query $\to$ Cond Logit & 0.7827 & 0.8793 & 0.7337 & 0.8392 & 7626 \\
MQ-C & Multi-Query $\to$ Logit $\alpha$=0.5 & 0.7843 & 0.8871 & 0.7356 & 0.8406 & 7945 \\
GATE-A & RRF + Hard Gate & 0.7802 & 0.9137 & 0.7308 & 0.8363 & 626 \\
GATE-D & RRF (vec=1.5) + Soft Gate & 0.7803 & 0.9086 & 0.7307 & 0.8362 & 1452 \\
O & Conditional Rerank $\alpha$=0.3 & 0.7792 & 0.9111 & 0.7294 & 0.8351 & 930 \\
S & Logit Blend + Cond $\alpha$=0.3 & 0.7792 & 0.9111 & 0.7294 & 0.8351 & 1066 \\
P & Full 9A+9B Pipeline & 0.7797 & 0.9099 & 0.7297 & 0.8354 & 1539 \\
RRF-D & Vector $\oplus$ Reranker RRF $k$=20 & 0.7798 & 0.9074 & 0.7300 & 0.8356 & 1502 \\
RRF-A & Vector $\oplus$ Reranker RRF $k$=60 & 0.7797 & 0.9074 & 0.7299 & 0.8355 & 1458 \\
GATE-B & RRF + Soft Gate & 0.7802 & 0.9074 & 0.7300 & 0.8357 & 1465 \\
GATE-C & RRF + Soft Gate thresh=0.12 & 0.7782 & 0.9049 & 0.7285 & 0.8345 & 1490 \\
RRF-B & Vector $\oplus$ Reranker RRF vec=1.5 & 0.7788 & 0.9062 & 0.7287 & 0.8349 & 1468 \\
MQ-C & Multi-Query $\to$ Logit $\alpha$=0.5 & 0.7843 & 0.8871 & 0.7356 & 0.8406 & 7945 \\
T & Logit Blend $\alpha$=0.3 & 0.7756 & 0.9137 & 0.7267 & 0.8323 & 1624 \\
M & Logit Blend $\alpha$=0.3 & 0.7756 & 0.9137 & 0.7267 & 0.8323 & 1525 \\
RRF-C & Vector $\oplus$ Reranker RRF rerank=1.5 & 0.7770 & 0.9037 & 0.7272 & 0.8333 & 1454 \\
A & Vector-Only & 0.7709 & 0.9037 & 0.7231 & 0.8293 & 516 \\
K & Vector $\to$ Adaptive MMR & 0.7622 & 0.8890 & 0.7156 & 0.8224 & 699 \\
MQ-A & Multi-Query $\to$ Vector-Only & 0.7619 & 0.8615 & 0.7150 & 0.8218 & 7550 \\
L & Full Adaptive Pipeline & 0.7515 & 0.8690 & 0.7035 & 0.8107 & 2057 \\
D & Hybrid + Reranker & 0.7525 & 0.8516 & 0.7047 & 0.8120 & 1486 \\
G & Full Pipeline & 0.7525 & 0.8516 & 0.7047 & 0.8120 & 1628 \\
I & Adaptive Hybrid & 0.7557 & 0.8641 & 0.7075 & 0.8148 & 1069 \\
H & Vector $\to$ Reranker & 0.7395 & 0.8340 & 0.6929 & 0.7990 & 1543 \\
R & Vector $\to$ Pure Logit $\alpha$=0 & 0.7395 & 0.8340 & 0.6929 & 0.7990 & 2381 \\
F & Hybrid + HyDE & 0.7278 & 0.8217 & 0.6809 & 0.7864 & 8828 \\
J & Reranker-as-Fusioner & 0.6993 & 0.7902 & 0.6527 & 0.7562 & 2113 \\
C & Hybrid RRF & 0.7307 & 0.8315 & 0.6838 & 0.7891 & 950 \\
E & Hybrid + MMR & 0.7113 & 0.8019 & 0.6659 & 0.7705 & 1179 \\
B & FTS-Only & 0.6587 & 0.7663 & 0.6159 & 0.7149 & 588 \\
\end{longtable}

\subsection{FiQA (648 queries)}

\begin{longtable}{llcccc}
\toprule
\textbf{ID} & \textbf{Configuration} & \textbf{NDCG@10} & \textbf{Recall@10} & \textbf{MAP@10} & \textbf{p50(ms)} \\
\midrule
\endfirsthead
\multicolumn{6}{l}{\emph{(continued)}} \\
\toprule
\textbf{ID} & \textbf{Configuration} & \textbf{NDCG@10} & \textbf{Recall@10} & \textbf{MAP@10} & \textbf{p50(ms)} \\
\midrule
\endhead
\bottomrule
\endlastfoot
U & Logit Blend $\alpha$=0.4 & 0.5526 & 0.6654 & 0.4311 & 1506 \\
V & Logit Blend $\alpha$=0.6 & 0.5526 & 0.6639 & 0.4309 & 1457 \\
N & Logit Blend $\alpha$=0.5 & 0.5526 & 0.6654 & 0.4309 & 1466 \\
GATE-A & RRF + Hard Gate & 0.5479 & 0.6543 & 0.4268 & 626 \\
A & Vector-Only & 0.5469 & 0.6512 & 0.4257 & 516 \\
H & Vector $\to$ Reranker & 0.5481 & 0.6481 & 0.4266 & 1543 \\
GATE-D & RRF (vec=1.5) + Soft Gate & 0.5409 & 0.6395 & 0.4207 & 1452 \\
GATE-B & RRF + Soft Gate & 0.5369 & 0.6327 & 0.4172 & 1465 \\
GATE-C & RRF + Soft Gate thresh=0.12 & 0.5371 & 0.6319 & 0.4173 & 1490 \\
RRF-D & Vector $\oplus$ Reranker RRF $k$=20 & 0.5479 & 0.6528 & 0.4265 & 1502 \\
RRF-A & Vector $\oplus$ Reranker RRF $k$=60 & 0.5479 & 0.6528 & 0.4265 & 1458 \\
RRF-B & Vector $\oplus$ Reranker RRF vec=1.5 & 0.5465 & 0.6497 & 0.4252 & 1468 \\
I & Adaptive Hybrid & 0.5045 & 0.5941 & 0.3913 & 1069 \\
C & Hybrid RRF & 0.4364 & 0.5077 & 0.3373 & 950 \\
D & Hybrid + Reranker & 0.4513 & 0.5173 & 0.3488 & 1486 \\
G & Full Pipeline & 0.4513 & 0.5173 & 0.3488 & 1628 \\
F & Hybrid + HyDE & 0.4196 & 0.4753 & 0.3234 & 8828 \\
B & FTS-Only & 0.2421 & 0.2669 & 0.1816 & 588 \\
\end{longtable}

\subsection{NFCorpus (323 queries)}

\begin{longtable}{llcccc}
\toprule
\textbf{ID} & \textbf{Configuration} & \textbf{NDCG@10} & \textbf{Recall@10} & \textbf{MAP@10} & \textbf{p50(ms)} \\
\midrule
\endfirsthead
\multicolumn{6}{l}{\emph{(continued)}} \\
\toprule
\textbf{ID} & \textbf{Configuration} & \textbf{NDCG@10} & \textbf{Recall@10} & \textbf{MAP@10} & \textbf{p50(ms)} \\
\midrule
\endhead
\bottomrule
\endlastfoot
A & Vector-Only & 0.4443 & 0.2212 & 0.2373 & 516 \\
I & Adaptive Hybrid & 0.4356 & 0.2169 & 0.2326 & 1069 \\
MQ-A & Multi-Query $\to$ Vector & 0.4341 & 0.2126 & 0.2318 & 7550 \\
K & Vector $\to$ Adaptive MMR & 0.4277 & 0.2042 & 0.2274 & 699 \\
MQ-C & Multi-Query $\to$ Logit $\alpha$=0.5 & 0.4292 & 0.2129 & 0.2291 & 7945 \\
U & Logit Blend $\alpha$=0.4 & 0.4344 & 0.2139 & 0.2317 & 1506 \\
RRF-B & Vector $\oplus$ Reranker RRF vec=1.5 & 0.4264 & 0.2068 & 0.2271 & 1468 \\
H & Vector $\to$ Reranker & 0.4256 & 0.2050 & 0.2263 & 1543 \\
GATE-A & RRF + Hard Gate & 0.4256 & 0.2050 & 0.2263 & 626 \\
RRF-D & Vector $\oplus$ Reranker RRF $k$=20 & 0.4255 & 0.2051 & 0.2262 & 1502 \\
RRF-A & Vector $\oplus$ Reranker RRF $k$=60 & 0.4256 & 0.2050 & 0.2263 & 1458 \\
L & Full Adaptive Pipeline & 0.4290 & 0.2086 & 0.2285 & 2057 \\
MQ-D & Multi-Query $\to$ Cond Logit & 0.4273 & 0.2096 & 0.2280 & 7626 \\
RRF-C & Vector $\oplus$ Reranker RRF rerank=1.5 & 0.4236 & 0.2033 & 0.2254 & 1454 \\
MQ-B & Multi-Query $\to$ Logit $\alpha$=0.4 & 0.4254 & 0.2096 & 0.2269 & 8690 \\
F & Hybrid + HyDE & 0.4171 & 0.2018 & 0.2220 & 8828 \\
C & Hybrid RRF & 0.4142 & 0.1965 & 0.2198 & 950 \\
G & Full Pipeline & 0.4105 & 0.1944 & 0.2180 & 1628 \\
D & Hybrid + Reranker & 0.4113 & 0.1947 & 0.2184 & 1486 \\
E & Hybrid + MMR & 0.4044 & 0.1893 & 0.2145 & 1179 \\
J & Reranker-as-Fusioner & 0.4072 & 0.1920 & 0.2162 & 2113 \\
GATE-D & RRF (vec=1.5) + Soft Gate & 0.4006 & 0.1876 & 0.2127 & 1452 \\
GATE-B & RRF + Soft Gate & 0.3996 & 0.1865 & 0.2118 & 1465 \\
GATE-C & RRF + Soft Gate thresh=0.12 & 0.3992 & 0.1859 & 0.2116 & 1490 \\
B & FTS-Only & 0.3146 & 0.1348 & 0.1648 & 588 \\
\end{longtable}

% ══════════════════════════════════════════════════════════════════════════════
\section{Pipeline Implementation Details}
\label{sec:appendix-pipeline}
% ══════════════════════════════════════════════════════════════════════════════

The following excerpts from \texttt{src/auto/recall.ts} show the actual implementation of key pipeline stages.

\subsection{Query Building (Lines 82--95)}

\begin{lstlisting}[language=bash, caption={Query construction from recent messages}]
function buildQuery(messages: unknown[], count: number): string {
  const lastN = messages.slice(-count);
  const parts: string[] = [];
  for (const msg of lastN) {
    const content = (msg as {content?: string}).content;
    if (typeof content === "string") {
      parts.push(content);
    }
  }
  return parts.join(" ").slice(0, 1000);
}
\end{lstlisting}

The query is built from the last $n_q$ agent messages (default: 2), truncated to 1000 characters. Previously injected memory blocks are stripped via \texttt{cleanQueryText()} to prevent querying for memories the system already injected.

\subsection{HyDE Integration (Lines 63--90)}

\begin{lstlisting}[language=bash, caption={HyDE: Hypothetical document generation}]
const hydeConfig = deps.hyde;
if (hydeConfig?.enabled) {
  console.log(`[recall] HyDE: generating hypothetical document...`);
  const hypothetical = await generateHypotheticalDocument(queryText, hydeConfig);
  if (hypothetical) {
    // HyDE: embed the hypothetical as a DOCUMENT (no instruction prefix).
    // The hypothetical is a pseudo-document and must land in document embedding
    // space, not query space. Per Gao et al. 2022.
    queryVector = await embed.embedDocument(hypothetical);
  } else {
    // HyDE failed (timeout, too short, etc.)
    queryVector = await embed.embedQuery(queryText);
  }
} else {
  queryVector = await embed.embedQuery(queryText);
}
\end{lstlisting}

Note the critical distinction: \texttt{embedDocument()} not \texttt{embedQuery()}. The hypothetical must land in document embedding space, with no instruction prefix. This is the correct implementation per Gao et al.\ (2023).

